---
title: "Problem Set 2^{K-P} Designs"
output: 
    html_document:
      number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Quarter Fraction Design

*This problem is adapted from Montgomery (2017) Problem 8.23*

A researcher is conducting an experiment using Monte Carlo simulation of an inventory system.  There are five factors: $A$, $B$, $C$, $D$, and $E$ that relate to the design parameters of the system.  The response is measured in cost.  The researcher could only do 8 runs using $I = ABD = BCE$ as the defining relations.  The results by treatment combination name are located here: <a href = '/_Chapter4_ProblemSets/Montgomery_8_23.csv'> Montgomery_8_23.csv </a>.  

## Understand the Design and the Data

1. Write out the alias structure for this design (only up to third order interactions is necessary):

```{r, message = F}
library(tidyverse)
# We can do this manually, but FrF2 is a useful function for this:

myDesign <- FrF2::FrF2(nfactors = 5, # We have 5 factors
                       nruns = 8, # We have a 2^(5-2) = 8 design, so 8 runs
                       generators = c('AB', 'BC'), # Recall the generators for FrF2 are odd, but 
                       # correspond to the last P letters of your design, i.e., D = AB, and E = BC
                       randomize = F, # We don't want to randomize the design
                       alias.info = 3 # we want to see up to third order interactions
                       )

attr(myDesign, "design.info")$aliased
# Here we can see our alias structure pretty clearly, note that this design is Res III
```

2. What is the resolution of this design?

```
This is a resolution III design as none of the main factors are aliased with each other, 
but are aliased with some of the 2nd order factors.  
```

3. Join the data given with the design in an appropriate way.  (Hint: You may find it helpful to figure a way to write the treatment combination names in the design and then do a join (e.g. `left_join` from dplyr to combine your data and design)).

```{r}
# Read the data
myData <- read.csv('Montgomery_8_23.csv', header = T, sep = ',')

# Consider our design:
myDesign
# We know that if a column has a positive number in it we use the lower case name of the column
# to codify the treatment combination name.
# In so far as I know, there isn't a useful functio for this, but we can do some dplyr-ing and make TCs:
# Basically we do this as a series of ifelse() functions and pasting with no seeparation
myDesign <- myDesign %>% mutate(TC = ifelse(A == '1', 'a', '')) %>% 
  mutate(TC = paste(TC, ifelse(B == '1', 'b', ''), sep = '')) %>% 
  mutate(TC = paste(TC, ifelse(C == '1', 'c', ''), sep = '')) %>% 
  mutate(TC = paste(TC, ifelse(D == '1', 'd', ''), sep = '')) %>% 
  mutate(TC = paste(TC, ifelse(E == '1', 'e', ''), sep = '')) %>% 
  mutate(TC = ifelse(TC == '', '(1)', TC))
#We can check our work
myDesign

# We can then join our data using left_join
myDF <- left_join(myDesign, myData)
# Note that left_join by default joins on matching column names
# and it will coerce data types to match if it can
myDF
```

## Visualize the Data

1. Visualize the main effects of the data. Comment on what you see.

```{r}
ggpubr::ggarrange(
  ggplot(myDF, aes(x = A, y = Cost)) + geom_boxplot(),
  ggplot(myDF, aes(x = B, y = Cost)) + geom_boxplot(),
  ggplot(myDF, aes(x = C, y = Cost)) + geom_boxplot(),
  ggplot(myDF, aes(x = D, y = Cost)) + geom_boxplot(),
  ggplot(myDF, aes(x = E, y = Cost)) + geom_boxplot()
)

# Recall, each of these boxes is just four data points.
# It looks as if A and B are significant, the others may not be.
```

## Effects

Estimate the effects of all main factors and any 2nd level interactions that are not aliased with main factors.

```{r}
# Recall that FrF2 produces levels as factors, not integers
# There may be an easier way to do this, but this works
myDF <- myDF %>% 
  mutate(A = as.integer(as.character(A))) %>% 
  mutate(B = as.integer(as.character(B))) %>% 
  mutate(C = as.integer(as.character(C))) %>% 
  mutate(D = as.integer(as.character(D))) %>% 
  mutate(E = as.integer(as.character(E)))

Effect.A <- (1/(2^(5-2-1)))*sum(myDF$A * myDF$Cost)
Effect.B <- (1/(2^(5-2-1)))*sum(myDF$B * myDF$Cost)
Effect.C <- (1/(2^(5-2-1)))*sum(myDF$C * myDF$Cost)
Effect.D <- (1/(2^(5-2-1)))*sum(myDF$D * myDF$Cost)
Effect.E <- (1/(2^(5-2-1)))*sum(myDF$E * myDF$Cost)
Effect.AC <- (1/(2^(5-2-1)))*sum(myDF$A * myDF$C * myDF$Cost)
Effect.AE <- (1/(2^(5-2-1)))*sum(myDF$A * myDF$E * myDF$Cost)

Effects <- data.frame(
  Factor.Name = c('A', 'B', 'C', 'D', 'E', 'AC', 'AE'),
  Effect = c(Effect.A, Effect.B, Effect.C, Effect.D, Effect.E, Effect.AC, Effect.AE)
)

Effects
```

2. Based on this, what factors do you consider significant?

```
It appears that A and B are almost certainly significant.
```

## Second Run

You were able to do a second run of 8 factors and got the following results:

```{r}
second.run <- data.frame(
  TC = c('ade', 'e', 'ab', 'bd', 'acd', 'c', 'abce', 'bcde'),
  Cost = c(136, 93, 187, 153, 139, 99, 191, 150)
)
```

You got this design by using the following generators (i.e., an alternate fraction): $D = -AB$ and $E = BC$.

Combine these results with your previous results:

```{r}
# We can choose to do this one of two ways.
# The first is that we can generate the design from the above generators as such:
second.design <- FrF2::FrF2(nfactors = 5, nruns = 8, generators = c('-AB', 'BC'), alias.info = 3, randomized = F)
second.design

# We can then proceed in the same manner:
second.design <- second.design %>% mutate(TC = ifelse(A == '1', 'a', '')) %>% 
  mutate(TC = paste(TC, ifelse(B == '1', 'b', ''), sep = '')) %>% 
  mutate(TC = paste(TC, ifelse(C == '1', 'c', ''), sep = '')) %>% 
  mutate(TC = paste(TC, ifelse(D == '1', 'd', ''), sep = '')) %>% 
  mutate(TC = paste(TC, ifelse(E == '1', 'e', ''), sep = '')) %>% 
  mutate(TC = ifelse(TC == '', '(1)', TC))

# Join the tables
second.DF <- left_join(second.design, second.run)

second.DF <- second.DF %>% 
  mutate(A = as.integer(as.character(A))) %>% 
  mutate(B = as.integer(as.character(B))) %>% 
  mutate(C = as.integer(as.character(C))) %>% 
  mutate(D = as.integer(as.character(D))) %>% 
  mutate(E = as.integer(as.character(E)))

# Bind the two data frames:
final.design <- rbind(myDF, second.DF)
final.design
# We can then proceed to analyze our data as before.
# The one challenge of this, is our alias structure is not immediately clear

# We can alternatively think about what our two sets of generators told us
# In the first one, take all TCs where D = AB and E = BC
# In the second one, take all TCs where D = -AB and E = BC
# We now have all possibilities for D, so we only have E = BC as a generator
# Let's generate that design

final.design.generated <- FrF2::FrF2(nfactors = 5, nruns = 16, generators = 'BC', randomize = F, alias.info = 3)

# Let's compare with final.design
# First, convert all of A,B, C, D, and E to factors:
final.design <- final.design %>% mutate(A = as.factor(A), B = as.factor(B), C = as.factor(C), D = as.factor(D), E = as.factor(E))
# Second, we can join on the generated design to see if we have every TC
results <- left_join(final.design.generated, final.design)
results
# It looks like this worked and is the correct understanding
# This is convenient, because we now have a design with our alias structure - final.design.generated
# and also a dataframe with all of our data, results, etc.
```

## Alias Structure

Explain the alias structure of the combined data and design.  What is its resolution?

```{r}
attr(final.design.generated, "design.info")$aliased
# We see B, C, and E are aliased with second order effects
# A and D are only aliased with 4th or higher
# We see a similar break up of 2nd order aliasing
# It is still resolution III due to the aliasing of B, C, and E
# But we have much higher resolution, particularly on A and D
# Note AD and CE must only be aliased with higher order effects (4th +)
```

## Estimate Effects

Using the combined data sets, estimate the effects for all main effects and the following 2nd order interactions:
AB, AC, AD, AE, BD, CD, and DE.  

```{r}
# We can proceed as we typically do:
# We first have to convert our factor values to integers in `results`
results <- results %>% 
  mutate(A = as.integer(as.character(A))) %>% 
  mutate(B = as.integer(as.character(B))) %>% 
  mutate(C = as.integer(as.character(C))) %>% 
  mutate(D = as.integer(as.character(D))) %>% 
  mutate(E = as.integer(as.character(E)))


Effect.A <- (1/(2^(5-1-1)))*sum(results$A * results$Cost)
Effect.B <- (1/(2^(5-1-1)))*sum(results$B * results$Cost)
Effect.C <- (1/(2^(5-1-1)))*sum(results$C * results$Cost)
Effect.D <- (1/(2^(5-1-1)))*sum(results$D * results$Cost)
Effect.E <- (1/(2^(5-1-1)))*sum(results$E * results$Cost)
Effect.AB <- (1/(2^(5-1-1)))*sum(results$A * results$B * results$Cost)
Effect.AC <- (1/(2^(5-1-1)))*sum(results$A * results$C * results$Cost)
Effect.AD <- (1/(2^(5-1-1)))*sum(results$A * results$D * results$Cost)
Effect.AE <- (1/(2^(5-1-1)))*sum(results$A * results$E * results$Cost)
Effect.BD <- (1/(2^(5-1-1)))*sum(results$B * results$D * results$Cost)
Effect.CD <- (1/(2^(5-1-1)))*sum(results$C * results$D * results$Cost)
Effect.DE <- (1/(2^(5-1-1)))*sum(results$D * results$E * results$Cost)

combined.effects <- data.frame(
  Factor.Name = c('A', 'B', 'C', 'D', 'E', 'AB', 'AC', 'AD', 'AE', 'BD', 'CD', 'DE'),
  Effect = c(Effect.A, Effect.B, Effect.C, Effect.D, Effect.E, Effect.AB, Effect.AC, Effect.AD, Effect.AE, Effect.BD, Effect.CD, Effect.DE)
)

combined.effects
# we can further see that A and B are significant
# We may consider the AB interaction, though it would be useful 
# with further testing to clarify that 

# temp.points <- qqnorm(combined.effects$Effect)
# qqline(combined.effects$Effect)
# text(x = temp.points$x, y = temp.points$y + 5, labels = combined.effects$Factor.Name)
```

## Other Design Options

In the above questions, you analyzed a $2^{5-2}$ design, then conducted a sequential experiment with an alternate fraction of the first design, effectively giving you a $2^{5-1}$ design.  Is your resulting design the highest resolution choice for a $2^{5-1}$ design?  If so, why?  If not, what is a better option?

```{r}
# Our resulting design was Resolution III.
# With five factors and a half fractional design, we can get a resolution V design
# For example, we can create a design using FrF2:
sample.design <- FrF2::FrF2(nfactors = 5, nruns = 16, alias.info = 3, randomize = F)
# We can see its resolution here:
attr(sample.design, "design.info")$catlg.entry
# We can confirm this by seeing what is aliased with main, second, and third order interactions
# I'm not quite clear how to read the generator column from FrF2, but it's clearly I = ABCDE
attr(sample.design, 'design.info')$aliased
# Note FrF2 only shows up to third order interactions, so all of the main interactions are
# aliased with 4th order or higher
# the seconds order interactions are only aliased with third order
# this meets our requirement for resolution V
```