---
title: "Introduction to Factorial Designs Problem Set - 2^K Designs"
output: 
  html_document:
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


# $2^3$ Design

*Note, this is adapted from ยง6.10 of (Montgomery, 2017).*^[Note, this problem was initally a question about a machining tool, but the analysis is equally valid for simulation results]

You conducted a simulation to assess the impact of three factors, A, B, and C at a low and high level each.  You used a $2^3$ full factorial design.  You were able to do three replicates of each treatment combination and conducted the simulation in a random order so that you would meet all assumptions.  Your data is contained in <a href = '/_Chapter4_ProblemSets/montgomery_6_1.csv'> montgomery_6_1.csv </a>.

```{r}
# 
myData <- read.csv('montgomery_6_1.csv')
head(myData)
```

## Visualize the Data

Visualize the data using both boxplots and interaction plots.  Interpret your results.

```{r}
# Box plots
# Note, we need to change the Factors to factors (R data types), but I didn't want to yet
# also we need to format our data nicely

myTempData <- myData %>% mutate(A = as.factor(A), B = as.factor(B), C = as.factor(C)) %>% pivot_longer(cols = c(Rep.1, Rep.2, Rep.3), names_to = 'Rep', values_to = 'Response')

ggpubr::ggarrange(
  ggplot(myTempData) + 
    geom_boxplot(aes(x = A, y = Response )),
  ggplot(myTempData) + 
    geom_boxplot(aes(x = B, y = Response )),
  ggplot(myTempData) + 
    geom_boxplot(aes(x = C, y = Response )),
  nrow = 2, ncol = 2,
  labels = c('A', 'B', 'C')
)

# It appears that B has an impact; perhaps C, though that is less clear, almost certainly not A

# Interaction plots
ggpubr::ggarrange(
  ggplot(myTempData, aes(x = A, y = Response, color = B, group = B)) +
    stat_summary(fun = 'mean', geom = 'line') +
    stat_summary(fun = 'mean', geom = 'point'),
  ggplot(myTempData, aes(x = A, y = Response, color = C, group = C)) +
    stat_summary(fun = 'mean', geom = 'line') +
    stat_summary(fun = 'mean', geom = 'point'),
  ggplot(myTempData, aes(x = B, y = Response, color = C, group = C)) +
    stat_summary(fun = 'mean', geom = 'line') +
    stat_summary(fun = 'mean', geom = 'point'),
  ncol = 2, nrow = 2, labels = c('AB', 'AC', 'BC')
)

# We see a pretty clear interaction between A and C, perhaps a bit less so between B and C, and not really between A and B
```

## Estimate the Effects

Estimate the effects using contrasts.

```{r}
myTempData <- myData %>% 
  mutate(AB = A*B, AC = A*C, BC = B*C, ABC = A*B*C, Response.Sum = Rep.1 + Rep.2 + Rep.3)

Effect.A <- sum(myTempData$A * myTempData$Response.Sum) / (3*2^(3-1))
Effect.B <- sum(myTempData$B * myTempData$Response.Sum) / (3*2^(3-1))
Effect.C <- sum(myTempData$C * myTempData$Response.Sum) / (3*2^(3-1))
Effect.AB <- sum(myTempData$AB * myTempData$Response.Sum) / (3*2^(3-1))
Effect.AC <- sum(myTempData$AC * myTempData$Response.Sum) / (3*2^(3-1))
Effect.BC <- sum(myTempData$BC * myTempData$Response.Sum) / (3*2^(3-1))
Effect.ABC <- sum(myTempData$ABC * myTempData$Response.Sum) / (3*2^(3-1))

data.frame(Factor = c('A', 'B', 'C', 'AB', 'AC', 'BC', 'ABC'), 
           Effect = c(Effect.A, Effect.B, Effect.C, Effect.AB, Effect.AC, Effect.BC, Effect.ABC))

# We see that the effect of B appears to be the largest and clearly non-zero; B additionally
# We see our interactions of AC is significant, with perhaps a 2 way interactino of BC and 3-way with ABC
```

## Conduct ANOVA

Conduct a 3-factor ANOVA for the data set and assess your results.

```{r}
# Format and select data usefully
myTempData <- myData %>% 
  mutate(A = as.factor(A), B = as.factor(B), C = as.factor(C)) %>% 
  pivot_longer(cols = c(Rep.1, Rep.2, Rep.3), names_to = 'Rep', values_to = 'Response') %>% 
  select(-c(TC, Rep))

# Conduct the ANOVA
myAOV <- aov(Response ~ (.)^3, data = myTempData)

# View the response
summary(myAOV)
# This confirms our graphical and effects analysis that B, C, and AC are all significant at alpha = .01

# Plot for assumptions
par(mfrow = c(2,2))
plot(myAOV)
# We appear to meet assumptions
shapiro.test(myAOV$residuals)
bartlett.test(Response ~ TC, data = myData %>% pivot_longer(cols = c(Rep.1, Rep.2, Rep.3), names_to = 'Rep', values_to = 'Response'))
```

## Bonus

Visualize this design in three dimensions.

```{r}
myTempData <- myData %>% 
  mutate(A = as.factor(A), B = as.factor(B), C = as.factor(C)) %>% 
  pivot_longer(cols = c(Rep.1, Rep.2, Rep.3), names_to = 'Rep', values_to = 'Response') 


# ggplot doesn't, insofar as i know have a good 3D plotting capability
# plotly is a nice library for this though
# https://plotly-r.com/d-charts.html has a good example for how to do 3D plots
library(plotly)

# We can look at the design geometrically
plot_ly(myTempData, x = ~A, y = ~B, z = ~C) %>% 
  add_text(text = ~TC) %>% 
  add_markers()

myZ.CPos <- as.matrix(
  myTempData %>% 
    filter(C == 1) %>% 
    group_by(A, B) %>% 
    summarize(Mean = mean(Response)) %>% 
    pivot_wider(A, names_from = B, values_from = Mean) %>%
    ungroup() %>% 
    select(-A)
)             

myZ.CNeg <- as.matrix(
  myTempData %>% 
    filter(C == -1) %>% 
    group_by(A, B) %>% 
    summarize(Mean = mean(Response)) %>% 
    pivot_wider(A, names_from = B, values_from = Mean) %>%
    ungroup() %>% 
    select(-A)
)             
              
# We can also view a three D surface with interactions, but these get really hard
# to interpret, so this is more for fun.

# I'm not good at plotly so there's absolutely a better way to do this
# when C is high we see the surface in red
# When C is low we see the surface in blue

plot_ly() %>% 
  add_surface(x = c(-1, 1), y = c(-1, 1), z = ~myZ.CPos, colorscale = list(c(0, 1), c('rgb(255, 0, 0', 'rgb(255, 0, 0)'))) %>% 
  add_surface(x = c(-1, 1), y = c(-1, 1), z = ~myZ.CNeg, colorscale = list(c(0, 1), c('rgb(0, 0, 255', 'rgb(0, 0, 255)'))) %>% 
  layout(scene = list(
    xaxis = list(title = 'A'),
    yaxis = list(title = 'B'),
    zaxis = list(title = 'Mean Response')
  ))

```


# Conceptual

## Combinatorics

For a 6 factor, full factorial ($2^6$) design, with factors A, B, C, D, E, F and levels -1 and 1:

1. How many treatment combinations are there?

```{r}
# there are 2^6
2^6
```

2. What is the total number of factors and interaction effects?

Recall that $\sum_{k = 0}^{n} \binom{n}{k} = 2^n$.  But we will not have a "no factor effect" so we can subtract one

```{r}
# The toal number of factors and interaction effects is the sum of 6 choose k where n ranges from 1 to 6.  
2^6 - 1
# There are 63

# We can confirm this manually if desired
choose(6, 1) + choose(6, 2) + choose(6, 3) + choose(6, 4) + choose(6, 5) + choose(6, 6)
```

3. What are the levels for each factor for the treatment combination:
    * `(1)`
    * `ab`
    * `c`
    * `def`

```{}
(1): A = B = C = D = E = F = -1
ab: A = B = 1, C = D = E = F = -1
c: C = 1, A = B = D = E = F = -1
def: A = B = C = -1, D = E = F = 1
```

4. What is the "product" of:
    * ABC and BCD?
    * AB, ACE, ADEF?
    
```{}
ABC*BCD = A(B^2)CD = ACD
AB*ACE*ADEF = (A^3)BCD(E^2)F = ABCDF
```