---
title: "Introduction to Factorial Designs Problem Set - 2^K Designs"
output: 
    html_document:
        number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


# $2^3$ Design

*Note, this is adapted from ยง6.10 of (Montgomery, 2017).*^[Note, this problem was initally a question about a machining tool, but the analysis is equally valid for simulation results]

You conducted a simulation to assess the impact of three factors, A, B, and C at a low and high level each.  You used a $2^3$ full factorial design.  You were able to do three replicates of each treatment combination and conducted the simulation in a random order so that you would meet all assumptions.  Your data is contained in <a href = '/_Chapter4_ProblemSets/montgomery_6_1.csv'> montgomery_6_1.csv </a>.

## Visualize the Data

Visualize the data using both boxplots and interaction plots.  Interpret your results.

```{r}
# Answer here
```

## Estimate the Effects

Estimate the effects using contrasts.

```{r}
# Answer here
```

## Conduct ANOVA

Conduct a 3-factor ANOVA for the data set and assess your results.

```{r}
# Answer here
```

## Bonus

Visualize this design in three dimensions.

```{r}
# Answer here
```


# Conceptual

## Combinatorics

For a 6 factor, full factorial ($2^6$) design, with factors A, B, C, D, E, F and levels -1 and 1:

1. How many treatment combinations are there?

```{r}
# Answer here
```

2. What is the total number of factors and interaction effects?

```{r}
# Answer here
```

3. What are the levels for each factor for the treatment combination:
    * `(1)`
    * `ab`
    * `c`
    * `def`

```{}
# Answer here
```

4. What is the "product" of:
    * ABC and BCD?
    * AB, ACE, ADEF?
    
```{}
# Answer here
```