---
title: "Introduction to Factorial Designs Problem Set - 2^K Designs"
output: 
    html_document:
      number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


# Half Fractional Question

*Note, this problem is adapted from Montgomery 2017, Question 8.10*

You have an experiment with five factors, A, B, C, D, and E.  You ran three replicates of each of 16 treatment combinations.  Your results are contained in <a href = '/_Chapter4_ProblemSets/montgomery_8_10.csv'> montgomery_8_10.csv </a>.  

## Read and Format the Data as Required

```{r}
#Answer here
```

## Understand the Design

1. Write out the alias structure of this design.

```{r}
#Answer here
```

```
#Answer here
```

2. What is the design generator?  What is the defining relation?  Is this a principal or alternate fraction?

```
#Answer here
```

## Analyze the Results

1. Visualize the main effects and interpret your results

```{r}
#Answer here
```

2. Visualize 2nd order interaction effects.

```{r}
#Answer here
```

3. Calculate the effect for each main and second order effect:

```{r}
#Answer here
```

4. Plot the effects on a normal probability plot.

```{r}
#Answer here
```

5. Based on your visualization and effects analysis, what do you conclude?

```
#Answer here
```

# Conceptual Question

If you are in a scenario in which you have many factors, each at two levels, but only a fixed number of possible runs, does it make more sense to:

* Do a full factorial experiment with only a single replicate of each treatment combination?
* Do a fractional factorial experiment with multiple replicates of the tested treatment combinations?

```
#Answer here
```
