---
title: "Introduction to R Problem Set"
author: "Your Name Here"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This problem set corresponds to the Introduction to *R* chapter in the course https://doe.dscoe.org. The solutions to the problem set use the packages shown below.

```{r}
library(tidyverse)
library(nycflights13)
```

# 3.1 Exercises

There are no exercises associated with this section.

# 3.2 Exercises

Complete all exercises for this section.

## 1

Run ggplot(data = mpg). What do you see?

```{r}
ggplot(data = mpg)

# produces a blank plot because we haven't given it a geometry yet.
```

## 2 

How many rows are in mpg? How many columns?

```{r}
# method 1
dim(mpg)
# 234 rows, 11 columns

# method 2
nrow(mpg)
ncol(mpg)

# method 3 - look at the summary at the bottom of the dataframe
mpg
```

## 3

What does the `drv` variable describe? Read the help for `?mpg` to find out.

```{r}
# not run
# ?mpg

# f = front-wheel drive, r = rear wheel drive, 4 = 4wd
```

## 4

Make a scatter plot of `hwy` vs. `cyl`.

```{r}
ggplot(mpg, aes(x=hwy, y=cyl)) +
  geom_point()
```

## 5

What happens if you make a scatter plot of `class` vs `drv`? Why is the plot not useful?

```{r}
ggplot(mpg, aes(x=class, y=drv)) +
  geom_point()

# class and drv are both unordered categorical variables, so "increasing" 
# values on the x and y axes have no meaning. Additionally, there are many 
# points plotted on top of each other, so you don't know how many points 
# are plotted for each pair of values.
```

# 3.3 Exercises

Complete all exercises for this section.

## 1

What’s gone wrong with this code? Why are the points not blue? `ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))`

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")

# color = "blue" needs to be outside of the aesthetic call. Using 
# color inside the aesthetic mapping assigns colors based on a third variable.
```

  
## 2

Which variables in mpg are categorical? Which variables are continuous? (Hint: type `?mpg` to read the documentation for the dataset). How can you see this information when you run `mpg`?

```{r}
str(mpg)

# continuous: displ
# integer: year, cyl, cty, hwy
# categorical: the rest of the variables

# or
# head(mpg)
```


## 3

Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = displ))

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = displ))

# not run
#ggplot(data = mpg) + 
#  geom_point(mapping = aes(x = displ, y = hwy, shape = displ))
# Error: A continuous variable can not be mapped to shape
```


## 4

What happens if you map the same variable to multiple aesthetics?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = fl, size = fl, shape = fl))

# Using the same variable doesn't add dimensions to the plot.
# It repeats the same information.
```

## 5

What does the stroke aesthetic do? What shapes does it work with? (Hint: use `?geom_point`)

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), shape = 21, stroke = 0.5)

# Works with shapes the have a border - it changes the thickness of the border.
```

## 6

What happens if you map an aesthetic to something other than a variable name, like `aes(colour = displ < 5)`? Note, you’ll also need to specify `x` and `y`.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = displ < 5))
```

# 3.4 Exercises

There are no exercises for this section.

# 3.5 Exercises

Complete exercises 1-4.

## 1

What happens if you facet on a continuous variable?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(~displ)
```

## 2

What do the empty cells in plot with `facet_grid(drv ~ cyl)` mean? How do they relate to this plot?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)

# It means there are no x and y values associated with that combination 
# of drv and cyl. For example, there are no cars in the dataset with 4WD 
# and 5 cylinders.
```

## 3

What plots does the following code make? `ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy)) + facet_grid(drv ~ .)` What does `.` do?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ .)

# `.` is a wildcard for including the combination of all other variables.
```

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(. ~ cyl)
```
  
## 4

Take the first faceted plot in this section:

`ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy)) + facet_wrap(~ class, nrow = 2)`

What are the advantages to using faceting instead of the colour aesthetic? What are the disadvantages? How might the balance change if you had a larger dataset?

```{r}
# There is no single correct answer. Faceting can help de-clutter a plot,
# but then it also makes it more difficult to directly compare two 
# categories. 
```

# 3.6 Exercises

Complete all exercises for this section.

## 1

What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?

```{r}
# line: geom_line or geom_path
# boxplot: geom_boxplot
# histogram: geom_histogram
# area chart: geom_area
```

## 2

Run this code in your head and predict what the output will look like. Then, run the code in *R* and check your predictions.

`ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + geom_point() + geom_smooth(se = FALSE)`

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() + 
  geom_smooth(se = FALSE)
```

## 3

What does `show.legend = FALSE` do? What happens if you remove it? Why do you think I used it earlier in the chapter?

```{r}
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv),
    show.legend = FALSE
  )
# show.legend = FALSE removes the plot legend.

ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv)
  )
```

## 4

What does the `se` argument to `geom_smooth()` do? Note: we will cover smoothers in detail in the non-parametric regression chapter of the DoE tutorial.

```{r}
# Displays a confidence interval for the smooth line (controlled with `level`).
```

## 5

Will these two graphs look different? Why/why not?

`ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + geom_point() + geom_smooth()`

`ggplot() + geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))`

```{r}
# They will look the same because both geom mappings in the second plot 
# are the same as the global mapping in the first plot.

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

## 6

Recreate the *R* code necessary to generate the following graphs (see R graphs in **R for Data Science** book)..

```{r}
ggplot(mpg, aes(x=displ, y=hwy)) +
  geom_point() +
  geom_smooth(se=FALSE)
```

```{r}
ggplot(mpg, aes(x=displ, y=hwy)) +
  geom_point() +
  geom_smooth(aes(group = drv), se=FALSE)
```

```{r}
ggplot(mpg, aes(x=displ, y=hwy)) +
  geom_point(aes(color=drv)) +
  geom_smooth(aes(color=drv), se=FALSE)
```

```{r}
ggplot(mpg, aes(x=displ, y=hwy)) +
  geom_point(aes(color=drv)) +
  geom_smooth(se=FALSE)
```

```{r}
ggplot(mpg, aes(x=displ, y=hwy)) +
  geom_point(aes(color=drv)) +
  geom_smooth(aes(group=drv, linetype=drv), se=FALSE)
```

```{r}
ggplot(mpg, aes(x=displ, y=hwy)) +
  geom_point(aes(color=drv)) +
  geom_point(shape = 21, stroke = 2, color='white')
```

# 3.7 Exercises

These exercises are optional, and no solutions are provided. 

# 3.8 Exercises

Complete exercises 1, 2, and 4 only. 

## 1

What is the problem with this plot? How could you improve it?

`ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + geom_point()`

```{r}
# Add jitter.
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_point(position = "jitter")

# alternatively:
#ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
#  geom_jitter()
```

## 2

What parameters to geom_jitter() control the amount of jittering?

```{r}
# width = 
# height =
```

## 4

What’s the default position adjustment for `geom_boxplot()`? Create a visualisation of the `mpg` dataset that demonstrates it.

```{r}
# Default position adjustment is dodge.
# Otherwise we would have superimposed boxplots.

ggplot(mpg, aes(x=class, y=hwy)) +
  geom_boxplot()
```

# 5.1

There are no exercises for this section.

# 5.2 Exercises

Complete exercises 1-3.

## 1 Find all flights that:

1. Had an arrival delay of two or more hours.

```{r}
flights = nycflights13::flights

flights %>% filter(arr_delay >= 120)
```

2. Flew to Houston (IAH or HOU).

```{r}
flights %>% filter(dest %in% c('IAH', 'HOU'))
```

3. Were operated by United, American, or Delta.

```{r}
flights %>% filter(carrier %in% c('UA', 'AA', 'DL'))
```

4. Departed in summer (July, August, and September).

```{r}
flights %>% filter(month %in% 7:9)
```

5. Arrived more than two hours late, but didn’t leave late.

```{r}
flights %>% filter(arr_delay > 120 & dep_delay <= 0)
```

6. Were delayed by at least an hour, but made up over 30 minutes in flight.

```{r}
flights %>% filter(dep_delay >= 60 & arr_delay < dep_delay - 30) 
# over 30 minutes means <, not <=
```

7. Departed between midnight and 6am (inclusive).

```{r}
flights %>% filter(dep_time == 2400 | dep_time <= 600)
```

## 2

Another useful dplyr filtering helper is `between()`. What does it do? Can you use it to simplify the code needed to answer the previous challenges?

```{r}
# 4 done again. Others follow the same format.
flights %>% filter(between(month, 7, 9))
```

## 3

How many flights have a missing `dep_time`? What other variables are missing? What might these rows represent?

```{r}
flights %>% filter(is.na(dep_time))
# 8,255 flights with missing dep_time. 
# Other missing variables: arrival and departure info. 
# Probably are canceled flights.
```

# 5.3 Exercises

Complete all exercises.

## 1

How could you use `arrange()` to sort all missing values to the start? (Hint: use `is.na()`).

```{r}
flights %>% arrange(desc(is.na(dep_time)))
```

## 2

Sort flights to find the most delayed flights. Find the flights that left earliest.

```{r}
# most delayed
flights %>% arrange(desc(dep_delay))

# left earliest
flights %>% arrange(dep_delay)
```

## 3

Sort flights to find the fastest (highest speed) flights.

```{r}
flights %>% arrange(desc(distance/air_time*60)) 
```

## 4

Which flights travelled the farthest? Which travelled the shortest?

```{r}
# the farthest
flights %>% arrange(desc(distance))

# the shortest
flights %>% arrange(distance)
```

# 5.4 Exercises

Complete exercises 2-4.

## 2

What happens if you include the name of a variable multiple times in a `select()` call?

```{r}
flights %>% select(carrier, carrier)

# It only includes the column once.
```

## 3

What does the `one_of()` function do? Why might it be helpful in conjunction with this vector?

`vars <- c("year", "month", "day", "dep_delay", "arr_delay")`

Note: the `tidyselect` documentation states:

>`one_of()` is retired in favour of the more precise `any_of()` and `all_of()` selectors.

```{r}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")

flights %>% select(one_of(vars))
```

## 4

Does the result of running the following code surprise you? How do the select helpers deal with case by default? How can you change that default?

`select(flights, contains("TIME"))`

```{r}
select(flights, contains("TIME"))

select(flights, contains("TIME", ignore.case=FALSE))

# select helpers ignore case
```

# 5.5 Exercises

Complete exercises 1-3.

## 1

Currently `dep_time` and `sched_dep_time` are convenient to look at, but hard to compute with because they’re not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight.

```{r}
flights %>% mutate(dep_since_midnight = (dep_time %/% 100) * 60 + dep_time %% 100,
                   sched_dep_since_midnight = (sched_dep_time %/% 100) * 60 + sched_dep_time %% 100
                   ) %>%
  select(dep_time, sched_dep_time, dep_since_midnight, sched_dep_since_midnight)
```

## 2

Compare `air_time` with `arr_time - dep_time`. What do you expect to see? What do you see? What do you need to do to fix it?

```{r}
flights %>% 
  mutate(diff = ((arr_time %/% 100) * 60 + arr_time %% 100) - 
           ((dep_time %/% 100) * 60 + dep_time %% 100)) %>%
  select(arr_time, dep_time, diff)

# Need to account for changing time zones and taxi-ing times.
```

## 3

Compare `dep_time`, `sched_dep_time`, and `dep_delay`. How would you expect those three numbers to be related?

```{r}
flights %>% 
  mutate(diff = ((dep_time %/% 100) * 60 + dep_time %% 100) - 
           ((sched_dep_time %/% 100) * 60 + sched_dep_time %% 100)) %>%
  select(diff, dep_delay)
```

# 5.6 Exercises

Complete exercises 1, 2, and 4. 

## 1

For this exercise, just find one way of solving the problem.

A flight is 15 minutes early 50% of the time, and 15 minutes late 50% of the time.

```{r}
not_cancelled = flights %>% filter(!is.na(dep_delay), !is.na(arr_delay))

not_cancelled %>% 
  group_by(flight) %>%
  summarize(prop_early = mean(arr_delay <= -15),
            prop_late = mean(arr_delay >= 15)) %>%
  filter(prop_early == 0.5 & prop_late == 0.5)
```

A flight is always 10 minutes late.

```{r}
not_cancelled %>%
  group_by(flight) %>%
  summarize(prop_late = mean(arr_delay >= 10)) %>%
  filter(prop_late == 1.0)
```

A flight is 30 minutes early 50% of the time, and 30 minutes late 50% of the time.

```{r}
not_cancelled %>% 
  group_by(flight) %>%
  summarize(prop_early = mean(arr_delay <= -30),
            prop_late = mean(arr_delay >= 30)) %>%
  filter(prop_early == 0.5 & prop_late == 0.5)
```

99% of the time a flight is on time. 1% of the time it’s 2 hours late.

```{r}
not_cancelled %>% 
  group_by(flight) %>%
  summarize(prop_ontime = mean(dep_delay <= 0),
            prop_late = mean(dep_delay >= 120)) %>%
  filter(prop_ontime >= 0.99 & prop_late <= 0.01)

```

## 2

Come up with another approach that will give you the same output as `not_cancelled %>% count(dest)` and `not_cancelled %>% count(tailnum, wt = distance)` without using `count()`.

```{r}
# not_cancelled %>% count(dest)
not_cancelled %>%
  group_by(dest) %>%
  summarize(dest_sum = n())

# not_cancelled %>% count(tailnum, wt = distance)
not_cancelled %>% 
  group_by(tailnum) %>%
  summarize(sum_dist = sum(distance))
```

## 4

Look at the number of cancelled flights per day. Is there a pattern? Is the proportion of cancelled flights related to the average delay?

```{r}
# number of cancelled flights per day faceted by month
flights %>%
  group_by(year, month, day) %>%
  summarize(n_cancel = sum(is.na(dep_delay))) %>%
  ggplot() +
  geom_col(aes(x=day, y=n_cancel)) +
  facet_wrap(~month)

# days combined from each month
flights %>%
  group_by(day) %>%
  summarize(n_cancel = sum(is.na(dep_delay))) %>%
  ggplot() +
  geom_col(aes(x=day, y=n_cancel))

flights %>%
  group_by(day) %>%
  summarize(prop_cancel = mean(is.na(dep_delay)),
            ave_delay = mean(dep_delay[!is.na(dep_delay)])) %>%
  ggplot() +
  geom_point(aes(x=prop_cancel, y=ave_delay)) +
  geom_smooth(aes(x=prop_cancel, y=ave_delay))
```

# 5.7 Exercises

Complete exercises 2 and 3.

## 2

Which plane (`tailnum`) has the worst on-time record?

```{r}
# If we define "worst on-time record" as the percent of flights a tailnum arrives late:

not_cancelled %>%
  group_by(tailnum) %>%
  summarize(pct_late = mean(arr_delay>0)) %>%
  arrange(desc(pct_late))
```

## 3

What time of day should you fly if you want to avoid delays as much as possible?

```{r}
# Looking at it from the perspective of departure delay:

not_cancelled %>%
  mutate(time_of_day = case_when(
    sched_dep_time < 0600 ~ "early_morning",
    between(sched_dep_time, 0600, 1200) ~ "morning",
    between(sched_dep_time, 1200, 1800) ~ "afternoon",
    sched_dep_time > 1800 ~ "night")) %>%
  group_by(time_of_day) %>%
  summarize(pct_late = mean(dep_delay>0)) %>%
  arrange(pct_late)

# Alternatively, we could define "avoid delays" as arriving on time.
```
