---
title: "Introduction to R Problem Set"
author: "Your Name Here"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This problem set corresponds to the Introduction to *R* chapter in the course https://doe.dscoe.org. The solutions to the problem set use the packages shown below.

```{r}
library(tidyverse)
library(nycflights13)
```

# 3.1 Exercises

There are no exercises associated with this section.

# 3.2 Exercises

Complete all exercises for this section.

## 1

Run ggplot(data = mpg). What do you see?

```{r}
# Answer here.
```

## 2 

How many rows are in mpg? How many columns?

```{r}
# Answer here.
```

## 3

What does the `drv` variable describe? Read the help for `?mpg` to find out.

```{r}
# Answer here.
```

## 4

Make a scatter plot of `hwy` vs. `cyl`.

```{r}
# Answer here.
```

## 5

What happens if you make a scatter plot of `class` vs `drv`? Why is the plot not useful?

```{r}
# Answer here.
```

# 3.3 Exercises

Complete all exercises for this section.

## 1

What’s gone wrong with this code? Why are the points not blue? `ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))`

```{r}
# Answer here.
```

  
## 2

Which variables in mpg are categorical? Which variables are continuous? (Hint: type `?mpg` to read the documentation for the dataset). How can you see this information when you run `mpg`?

```{r}
# Answer here.
```


## 3

Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?

```{r message=FALSE, warning=FALSE}
# Answer here.
```


## 4

What happens if you map the same variable to multiple aesthetics?

```{r}
# Answer here.
```

## 5

What does the stroke aesthetic do? What shapes does it work with? (Hint: use `?geom_point`)

```{r}
# Answer here.
```

## 6

What happens if you map an aesthetic to something other than a variable name, like `aes(colour = displ < 5)`? Note, you’ll also need to specify `x` and `y`.

```{r}
# Answer here.
```

# 3.4 Exercises

There are no exercises for this section.

# 3.5 Exercises

Complete exercises 1-4.

## 1

What happens if you facet on a continuous variable?

```{r}
# Answer here.
```

## 2

What do the empty cells in plot with `facet_grid(drv ~ cyl)` mean? How do they relate to this plot?

```{r}
# Answer here.
```

## 3

What plots does the following code make? `ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy)) + facet_grid(drv ~ .)` What does `.` do?

```{r}
# Answer here.
```

```{r}
# Answer here.
```
  
## 4

Take the first faceted plot in this section:

`ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy)) + facet_wrap(~ class, nrow = 2)`

What are the advantages to using faceting instead of the colour aesthetic? What are the disadvantages? How might the balance change if you had a larger dataset?

```{r}
# Answer here.
```

# 3.6 Exercises

Complete all exercises for this section.

## 1

What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?

```{r}
# Answer here.
```

## 2

Run this code in your head and predict what the output will look like. Then, run the code in *R* and check your predictions.

`ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + geom_point() + geom_smooth(se = FALSE)`

```{r}
# Answer here.
```

## 3

What does `show.legend = FALSE` do? What happens if you remove it? Why do you think I used it earlier in the chapter?

```{r}
# Answer here.
```

## 4

What does the `se` argument to `geom_smooth()` do? Note: we will cover smoothers in detail in the non-parametric regression chapter of the DoE tutorial.

```{r}
# Answer here.
```

## 5

Will these two graphs look different? Why/why not?

`ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + geom_point() + geom_smooth()`

`ggplot() + geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))`

```{r}
# Answer here.
```

## 6

Recreate the *R* code necessary to generate the following graphs (see R graphs in **R for Data Science** book).

```{r}
# Answer here.
```

# 3.7 Exercises

These exercises are optional, and no solutions are provided. 

# 3.8 Exercises

Complete exercises 1, 2, and 4 only. 

## 1

What is the problem with this plot? How could you improve it?

`ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + geom_point()`

```{r}
# Answer here.
```

## 2

What parameters to `geom_jitter()` control the amount of jittering?

```{r}
# Answer here.
```

## 4

What’s the default position adjustment for `geom_boxplot()`? Create a visualisation of the `mpg` dataset that demonstrates it.

```{r}
# Answer here.
```

# 5.1

There are no exercises for this section.

# 5.2 Exercises

Complete exercises 1-3.

## 1 Find all flights that:

1. Had an arrival delay of two or more hours.

```{r}
# Answer here.
```

2. Flew to Houston (IAH or HOU).

```{r}
# Answer here.
```

3. Were operated by United, American, or Delta.

```{r}
# Answer here.
```

4. Departed in summer (July, August, and September).

```{r}
# Answer here.
```

5. Arrived more than two hours late, but didn’t leave late.

```{r}
# Answer here.
```

6. Were delayed by at least an hour, but made up over 30 minutes in flight.

```{r}
# Answer here.
```

7. Departed between midnight and 6am (inclusive).

```{r}
# Answer here.
```

## 2

Another useful dplyr filtering helper is `between()`. What does it do? Can you use it to simplify the code needed to answer the previous challenges?

```{r}
# Answer here.
```

## 3

How many flights have a missing `dep_time`? What other variables are missing? What might these rows represent?

```{r}
# Answer here.
```

# 5.3 Exercises

Complete all exercises.

## 1

How could you use `arrange()` to sort all missing values to the start? (Hint: use `is.na()`).

```{r}
# Answer here.
```

## 2

Sort flights to find the most delayed flights. Find the flights that left earliest.

```{r}
# Answer here.
```

## 3

Sort flights to find the fastest (highest speed) flights.

```{r}
# Answer here.
```

## 4

Which flights travelled the farthest? Which travelled the shortest?

```{r}
# Answer here.
```

# 5.4 Exercises

Complete exercises 2-4.

## 2

What happens if you include the name of a variable multiple times in a `select()` call?

```{r}
# Answer here.
```

## 3

What does the `one_of()` function do? Why might it be helpful in conjunction with this vector?

`vars <- c("year", "month", "day", "dep_delay", "arr_delay")`

Note: the `tidyselect` documentation states:

>`one_of()` is retired in favour of the more precise `any_of()` and `all_of()` selectors.

```{r}
# Answer here.
```

## 4

Does the result of running the following code surprise you? How do the select helpers deal with case by default? How can you change that default?

`select(flights, contains("TIME"))`

```{r}
# Answer here.
```

# 5.5 Exercises

Complete exercises 1-3.

## 1

Currently `dep_time` and `sched_dep_time` are convenient to look at, but hard to compute with because they’re not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight.

```{r}
# Answer here.
```

## 2

Compare `air_time` with `arr_time - dep_time`. What do you expect to see? What do you see? What do you need to do to fix it?

```{r}
# Answer here.
```

## 3

Compare `dep_time`, `sched_dep_time`, and `dep_delay`. How would you expect those three numbers to be related?

```{r}
# Answer here.
```

# 5.6 Exercises

Complete exercises 1, 2, and 4. 

## 1

For this exercise, just find one way of solving the problem.

A flight is 15 minutes early 50% of the time, and 15 minutes late 50% of the time.

```{r}
# Answer here.
```

A flight is always 10 minutes late.

```{r}
# Answer here.
```

A flight is 30 minutes early 50% of the time, and 30 minutes late 50% of the time.

```{r}
# Answer here.
```

99% of the time a flight is on time. 1% of the time it’s 2 hours late.

```{r}
# Answer here.
```

## 2

Come up with another approach that will give you the same output as `not_cancelled %>% count(dest)` and `not_cancelled %>% count(tailnum, wt = distance)` without using `count()`.

```{r}
# Answer here.
```

## 4

Look at the number of cancelled flights per day. Is there a pattern? Is the proportion of cancelled flights related to the average delay?

```{r}
# Answer here.
```

# 5.7 Exercises

Complete exercises 2 and 3.

## 2

Which plane (`tailnum`) has the worst on-time record?

```{r}
# Answer here.
```

## 3

What time of day should you fly if you want to avoid delays as much as possible?

```{r}
# Answer here.
```
