---
title: "Multi Factor ANOVA Problem Set"
author: 'Your Name Here'
output: 
  html_document:
    number_sections: T
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This problem set corresponds to the section on multi factor ANOVA in the course https://doe.dscoe.org.  

For this problem set, we will use the following packages:

```{r, warning = F, message = F}
library(tidyverse)
library(ggpubr) # Note, we haven't used it yet in the course, but its a convenient package for combining multiple plots
```


# Simulation

You are an analyst working on a top secret program and are given the file <a href = '/_Chapter3_1_ProblemSets/multi_factor.csv'> multi_factor.csv </a> to analyze.^[The reality is, this was data from an industrial application of silicon wafers, but who cares about that?]  The data shows the results for a simulation where four factors, A, B, C, and D are measured at a "low" (represented as -1) and "high" (represented as 1) level and their response is as indicated.  There was only one replicate for each treatment combination^[We have not yet introduced this term, but this is the combination of the levels of each factor for an experiment / simulation run.  For example, Factor A, Factor B, Factor C, Factor D all set to "-1" is one treatment combination.]

## Input the Data

Read the data and clean it as necessary.

```{r}
# Answer here
```

## Visualize the Data

Plot the data in a way that allows you to understand something about the effects (main or interaction) of the factors.  Interpret your results.

Note - we are dealing with five dimensional data (the Response and factors A, B, C, and D); there will be many ways to view this.

```{r}
# Answer here
```

## Conduct ANOVA

Conduct an four ANOVAs on this data:

* One where you only consider the main effects.
* One where you consider the main effects and all two-way interactions.
* One where you consider the above plus all three-way interactions.
* One where you consider the above plus the four-way interaction.

For each ANOVA, identify the factors you would consider significant.  Use $\alpha = .05$ for your analysis.

```{r}
# Answer here
```

What did you notice about:

* The p-values for the factors you initially considered significant.
* The degrees of freedom available for the residuals.

```{}
# Answer here
```

Why might there be a change?  (No need to explain this mathematically precisely, just conceptually).

```{}
# Answer here
```

## Pick a Model

Choose one model, explain why you chose this, and check its assumptions.

```{r}
# Answer here
```