# Non-Parametric Regression

```{r setup-9, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

As we've seen in the last few chapters, linear models can be successfully applied to many data sets. However, there may be times when even after transforming variables, your model clearly violates the assumptions of linear models. Alternatively, you may have evidence that there is a complex relationship between predictor and response that is difficult to capture through transformation. In these cases, non-parametric regression techniques offer alternative methods for modeling your data.

## Admin

For any errors associated with this section, please contact <a href="mailto:john.f.king1.mil@mail.mil">John King</a>.

```{r echo = F}
mySession <- sessionInfo()
```

This chapter was published using the following software:

* `r mySession$R.version$version.string`.
* On `r mySession$platform` running `r mySession$running`.
* Packages used in this chapter are explicitly shown in the code snippets.

## Non-Parametric ANOVA

In Chapter 4, we applied ANOVA to problems where we had a factor with three or more levels, and we wanted to test for a difference in response among the levels. One of the assumptions of parametric ANOVA is that the underlying data are normally distributed. If we have a data set that violates that assumption, as is often the case with counts (especially of relatively rare events), then we'll need to use a non-parametric method such as the Kruskal-Wallis (KW) test.

The setup for the KW test is as follows:

* Level 1: $X_{11}, X_{12}, ...,X_{1J_{1}} \sim F_{1}$
* Level 2: $X_{21}, X_{22}, ...,X_{2J_{2}} \sim F_{2}$
* ...
* Level I: $X_{I1}, X_{I2}, ...,X_{IJ_{I}} \sim F_{I}$

**Null hypothesis** $H_{o}: F_{1} = F_{2} = ... = F_{I}$

**Alternative hypothesis** $H_{a}:$ not $H{o}$ (i.e., at least two of the distributions are different).

The idea behind the KW test is to sort the data and use an observation's rank instead of the value itself. If we have a sample size $N = J_{1}, J_{2}, ... J_{I}$, we first rank the observations (the lowest value gets a 1, second lowest a 2, etc.). If there are ties, then use the mid-rank (the mean of the two ranks). Then, separate the samples into their respective levels and sum the ranks for each level. For example, if we have three levels:

* Level 1: $R_{11} + R_{12} + ... + R_{1J_{1}}=$ sum of ranks
* Level 2: $R_{21} + R_{22} + ... + R_{2J_{1}}=$ sum of ranks
* Level 3: $R_{31} + R_{32} + ... + R_{3J_{1}}=$ sum of ranks

Now we do the non-parametric equivalent of a sum of squares for treatment (SSTr) using these ranks. The KW test statistic takes the form:

$$K=\frac{12}{N(N+1)}\sum\limits_{i=1}^{I}{\frac{R^2_{i}}{J_{i}}-3(N+1)}$$

For hypothesis testing, we calculate a p-value where large values of K signal rejection. We do this by approximating K with a chi-square distribution having $I-1$ degrees of freedom. According to @devore2015 (p. 672), chi-square is a good approximation for K if:

1. If there are I = 3 treatments and each sample size is >= 6, or
2. If there are I > 3 treatments and each sample size is >= 5

Let's look at an example. Say we are a weapon manufacturer that produces rifles on three different assembly lines, and we want to know if there's a difference in the number of times a weapon jams. We select 10 random weapons from each assembly line and perform our weapon jamming test identically on all 30 weapons. I'll make up some dummy data for this situation by drawing random numbers from Poisson distributions with two different $\lambda$s.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
set.seed(42)

j = c(rpois(10, 10), rpois(10, 10), rpois(10, 15))
r = rank(j, ties.method="average")
l = paste("line", rep(1:3, each = 10), sep="")

tibble(
  jams1 = j[1:10],
  rank1 = r[1:10],
  jams2 = j[11:20],
  rank2 = r[11:20],
  jams3 = j[21:30],
  rank3 = r[21:30])

```

For this data, 

* $I = 3$
* $J_{1} = J_{2} = J_{3} = 10$
* $N = 10 + 10 + 10 = 30$
* $R_{1} =$ `r sum(r[1:10])`
* $R_{2} =$ `r sum(r[11:20])`
* $R_{3} =$ `r sum(r[21:30])`

Which gives the K statistic:

$$K=\frac{12}{30(31)} \left[\frac{120^2}{10} + \frac{131.5^2}{10} + \frac{213.5^2}{10}\right] - 3(31)$$


```{r}
K = 12/(30*31) * (sum(r[1:10])^2/10 + sum(r[11:20])^2/10 + sum(r[21:30])^2/10) - 3*31
print(paste("K =", K), quote=FALSE)
```

Then, compute an approximate p-value using the chi-square test with two degrees of freedom.

```{r}
1 - pchisq(K, df=2)
```

We therefore reject the null hypothesis at the $\alpha = 0.5$ test level. Of course, there's an *R* function  `kruskall.test()` so we don't have to do all that by hand.

```{r}
rifles = tibble(jams = j, line = l)

kruskal.test(jams ~ line, data = rifles)
```

### Multiple Comparisons

The KW test can also be used for multiple comparisons. Recall that we applied the Tukey Test to the parametric case, and it took into account that there is an increase in the probability of a Type I error when conducting multiple comparisons. The non-parametric equivalent is to combine the Bonferroni Method with the KW test. Consider the case where we conduct $m$ tests of the null hypothesis. We calculate the probability that at least one of the null hypotheses is rejected ($P(\bar{A})$) as follows:

$$P(\bar{A}) = 1-P(A) = 1-P(A_{1} \cap A_{2} \cap...\cap A_{m})$$

$$=1-P(A_{1})P(A_{2})\cdot\cdot\cdot P(A_{m})$$

$$=1-(1-\alpha)^{m}$$

So if our individual test level target is $\alpha=0.05$ and we conduct $m=10$ tests, then $P(\bar{A})=1-(1-0.05)^{10}=$ `r 1-(1-0.05)^10`. If we want to establish a **family-wide Type I error rate**, $\Psi=P(\bar{A})=0.05$, then the individual test levels should be:

$$\alpha=P(\bar{A}_{i})=1-(1-\Psi)^{1/m}$$

For example, if $\Psi=0.05$ and we again conduct $m=10$ tests, then $\alpha=1-(1-0.05)^{1/10}=$ `r 1-(1-0.05)^(1/10)`. The downfall of this approach is that the same data are used to test the collection of hypotheses, which violates the independence assumption. The Bonferroni Inequality saves us from this situation because it doesn't rely on the assumption of independence. Using the Bonferroni method, we simply calculate $\alpha=\Psi/m = 0.005$. Note that this is almost identical to the result when we assumed independence. Unfortunately, the method isn't perfect. As noted in <a href="https://www.stat.berkeley.edu/~mgoldman/Section0402.pdf">Section 2 of this paper</a> the Bonferroni method tends to be overly conservative, which increases the chance of a false negative. 

Using the `rifles` data from earlier, if we wanted to conduct all three pair-wise hypothesis tests, then $\alpha=0.05/3=0.01667$ for the individual KW tests.

The `dunn.test()` function from the aptly-named `dunn.test` package performs the multiple pair-wise tests and offers several methods for accounting for performing multiple tests, including Bonferroni. For example (and notice we get the KW test results also):

```{r}
dunn.test::dunn.test(rifles$jams, rifles$line, method="bonferroni")
```

According to the `dunn.test` documentation, the null hypothesis for the Dunn test is:

>The null hypothesis for each pairwise comparison is that the probability of observing a randomly selected value from the first group that is larger than a randomly selected value from the second group equals one half; this null hypothesis corresponds to that of the Wilcoxon-Mann-Whitney rank-sum test. Like the ranksum test, if the data can be assumed to be continuous, and the distributions are assumed identical except for a difference in location, Dunn's test may be understood as a test for median difference. 'dunn.test' accounts for tied ranks.

### Non-Parametric ANOVA Problem Set 

The problem set for this section is located <a href = '/_Chapter11_ProblemSets/KW_PS_Questions.html'>here</a>.

For your convenience, the R markdown version is <a href = '/_Chapter11_ProblemSets/kw_PS_Questions.Rmd'>here</a>.

The solutions are located <a href = '/_Chapter11_ProblemSets/KW_PS_Solutions.html'>here</a>.

## Generalized Additive Models

Recall that if there is a non-linear relationship between predictor and response, we can attempt to transform the predictor using a known function (log, reciprocal, polynomial, etc.) to improve the model structure and fit. What if the relationship is more complex and is not well captured with a known function? Generalized additive models may be used in these cases. 

Recall that a linear model takes the form:

$$y=\beta_{0}+\beta_{1}x_{1}+\beta_{2}x_{2}+...+\varepsilon$$

Additive models replace the linear terms (the $\beta$s) with flexible smoothing functions and take the form:

$$y=\beta_{0}+f_{1}(x_{1})+f_{2}(x_{2})+...+\varepsilon$$

There are many techniques and options for selecting the smoothing functions, but for this tutorial, we'll discuss two: locally weighted error sum of squares (lowess and also commonly abbreviated as loess) and smoothing splines. 

### Loess

For the theory behind loess smoothing, please read <a href="https://www.itl.nist.gov/div898/handbook/pmd/section1/pmd144.htm">this page</a> on the NIST website. This chapter will focus on implementing loess smoothing in *R*. 

All smoothers have a tuning parameter that controls how smooth the smoother is. The tuning parameter in loess is referred to as the span with larger values producing more smoothness. 

```{r}
library(tidyverse)
set.seed(42)

df = tibble(
  x = runif(100, 1.5, 5.5),
  y = sin(x*pi) + 2 + runif(100, 0, 0.5)
)

ex1.ls = loess(y~x, span=0.25, data=df)
ex2.ls = loess(y~x, span=0.5, data=df)
ex3.ls = loess(y~x, span=0.75, data=df)
xseq = seq(1.6, 5.4,length=100)

df = df %>% mutate(
  span25 = predict(ex1.ls, newdata=tibble(x=xseq)),
  span50 = predict(ex2.ls, newdata=tibble(x=xseq)),
  span75 = predict(ex3.ls, newdata=tibble(x=xseq))
  )

ggplot(df) +
  geom_point(aes(x, y)) +
  geom_line(aes(x=xseq, span25, linetype='span = 0.25'), color='red') +
  geom_line(aes(x=xseq, span50, linetype='span = 0.50'), color='red') +
  geom_line(aes(x=xseq, span75, linetype='span = 0.75'), color='red') +
  scale_linetype_manual(name="Legend", values=c(1,2,3)) +
  ggtitle("Loess Smoother Example") +
  theme_bw()
```

From this plot, a span of 0.75 provided too much smoothness, whereas the lower values of span we tested appear to be a better fit. Now let's apply this to the `airqaulity` data set from the previous chapter. Initially, we'll just consider the response(`Ozone`) and one predictor (`Solar.R`).

```{r message=FALSE, warning=FALSE}
aq1.ls = loess(Ozone ~ Solar.R, span=0.25, data=airquality)
aq2.ls = loess(Ozone ~ Solar.R, span=0.5, data=airquality)
aq3.ls = loess(Ozone ~ Solar.R, span=0.75, data=airquality)
srseq = seq(10, 330, length=nrow(airquality))

aq = airquality %>% mutate(
  span25 = predict(aq1.ls, newdata=tibble(Solar.R=srseq)),
  span50 = predict(aq2.ls, newdata=tibble(Solar.R=srseq)),
  span75 = predict(aq3.ls, newdata=tibble(Solar.R=srseq))
  )

ggplot(aq) +
  geom_point(aes(Solar.R, Ozone)) +
  geom_line(aes(x=srseq, span25, linetype='span = 0.25'), color='red') +
  geom_line(aes(x=srseq, span50, linetype='span = 0.50'), color='red') +
  geom_line(aes(x=srseq, span75, linetype='span = 0.75'), color='red') +
  scale_linetype_manual(name="Legend", values=c(1,2,3)) +
  ggtitle("Loess Smoother Example") +
  theme_bw()
```

Here we can see that the higher span values appear to provide a better fit. In this case, choosing a low span value would be akin to over fitting a linear model with too high of a degree of polynomial. We can repeat this process to determine appropriate values of span for the other predictors. 

Including loess smoothers in a GAM is as simple as including the non-linear terms within `lo()`. The `gam` package provides the needed functionality. The script below applies loess smoothers to three of the predictors and displays the model summary (note that the default value for span is 0.5).

```{r message=FALSE, warning=FALSE}
library(gam)

aq.gam = gam(Ozone ~ lo(Solar.R, span=0.75) + lo(Wind) + lo(Temp), data=airquality, na=na.gam.replace)
summary(aq.gam)
```

```{r, fig.width=10, fig.height=4}
par(mfrow=c(1,3))
plot(aq.gam, se=TRUE)
```

### Splines

Spline smoothing can be conceptualized by imagining that your task is to bend a strip of soft metal into a curved shape. One way to do this would be to place pegs on a board (referred to as "knots" in non-linear regression parlance) to control the bends, and then guide the strip of metal over and under the pegs. Mathematically, this is accomplished by combining cubic regression at each knot with calculus to smoothly join the individual bends. The tuning parameter in the `smooth.splines` function is `spar`.

```{r}
aq = aq %>% drop_na()

ss25 = smooth.spline(aq$Solar.R,aq$Ozone,spar=0.25)
ss50 = smooth.spline(aq$Solar.R,aq$Ozone,spar=0.5)
ss75 = smooth.spline(aq$Solar.R,aq$Ozone,spar=0.75)

ggplot() +
  geom_point(data=aq, aes(Solar.R, Ozone)) +
  geom_line(aes(x=ss25$x, ss25$y, linetype='spar = 0.25'), color='red') +
  geom_line(aes(x=ss50$x, ss50$y, linetype='spar = 0.50'), color='red') +
  geom_line(aes(x=ss75$x, ss75$y, linetype='spar = 0.75'), color='red') +
  scale_linetype_manual(name="Legend", values=c(1,2,3)) +
  ggtitle("Spline Smoother Example") +
  theme_bw()
```

### Cross Validation

Comparing the spline smoother plot to the one generated with loess smoothers, we can see that the two methods essentially accomplish the same thing. It's just a matter of finding the right amount of smoothness, which can be done through cross validation. The `fANCOVA` package contains a function `loess.aq()` that includes a criterion parameter that we can set to `gcv` for generalized cross validation, which is an approximation for leave-one-out cross-validation @hastie2008. Applying this function to the `airquality` data with `Solar.R` as the predictor and `Ozone` as the response, we can obtain a cross validated value for span.

```{r}
library(fANCOVA)

aq.solar.cv = loess.as(aq$Solar.R, aq$Ozone, criterion="gcv")
summary(aq.solar.cv)
```

`loess.as` also includes a plot method so we can visualize the loess smoother.

```{r}
loess.as(aq$Solar.R, aq$Ozone, criterion="gcv", plot=TRUE)
```

Cross validation is also built in to `smooth.spline()` and is set to generalized cross validation by default. Instead of specifying `spar` in the call to `smooth.spline()`, we just leave it out to invoke cross validation.

```{r}
aq.spl = smooth.spline(aq$Solar.R, aq$Ozone)
aq.spl
```

Plotting the cross validated spline smoother, we get a line that looks very similar to the lasso smoother.

```{r}
ggplot() +
  geom_point(data=aq, aes(Solar.R, Ozone)) +
  geom_line(aes(x=aq.spl$x, aq.spl$y), color='red') +
  ggtitle("CV Spline Smoother") +
  theme_bw()
```

### GAM Problem Set 

The problem set for this section is located <a href = '/_Chapter11_ProblemSets/Loess_PS_Questions.html'>here</a>.

For your convenience, the R markdown version is <a href = '/_Chapter11_ProblemSets/Loess_PS_Questions.Rmd'>here</a>.

The solutions are located <a href = '/_Chapter11_ProblemSets/Loess_PS_Solutions.html'>here</a>.

## Support Vector Machines

If you have already been introduced to support vector machines (SVM), chances are that the methodology was applied to a classification problem, which is referred to as support vector classification (SVC). Support vector regression (SVR) is closely related to SVC but is used for linear and non-linear regression problems. We'll begin with regression, and then move to classification.

### Support Vector Regression

SVR attempts to include as many data points as possible in the area between two lines. The following figure demonstrates this using dummy data with a linear relationship. The two parallel lines are the **margin**, and it's width is a hyperparameter $\varepsilon$ that we can tune. If you draw a line through one of the points that fall outside the margin so that it is perpendicular to the margin, you have a **support vector**. A **cost** is applied to each point that falls outside the margin, and minimizing the cost determines the slope of the margin. Cost is another tunable hyperparameter, which is sometimes represented as $1/\lambda$. Notice that unlike linear regression, if we were to add more points inside the margin, it would have no impact on the slope. SVR is also much less influence by outliers than linear regression. For the mathematical details behind SVR, refer to Section 12.3.6 in @hastie2008. 

```{r echo=FALSE}
set.seed(42)

df = tibble(
  x = seq(0,10,length=100),
  y = 1 + x + rnorm(100),
  sv = case_when(y > 2 + x | y < x - 1 ~ 'out',
                 y < 2 + x & y > x - 1 ~ 'in'))

M = matrix(c(1, -1, 1, 1), ncol=2, byrow= TRUE)
a = c(2, rowSums(df %>% select(x,y))[12])
xy = solve(M) %*% a

ggplot() +
  annotate("segment", x=df$x[12], xend=xy[2], y=df$y[12], yend = xy[1], color='blue') +
  annotate("text", x=1.5, y=7, label="Support\nVector", color='blue') +
  geom_segment(aes(x=1.5, xend=1.5, y=6.2, yend=4.5), color='blue', arrow=arrow(length = unit(0.1, "inches"))) +
  geom_abline(slope=1, intercept=-1) +
  geom_abline(slope=1, intercept=2) +
  geom_point(data = df, aes(x, y, color=sv)) +
  scale_color_manual(values = c('black', 'red')) +
  ggtitle("Support Vector Regression") +
  xlab("x") + ylab("y") +
  coord_fixed() +
  theme_bw() +
  theme(legend.position="none")

```

Choosing values for the hyperparameters $\varepsilon$ and $\lambda$ is once again done through cross validation. To do this in *R*, we'll use some functions from the `e1071` package (another option is the `LiblineaR` package). Before we get to cross validation, let's just look at how to build an SVR model. The syntax is the same as for linear models, we just replace `lm()` with `svm()`. Note that the function is not `svr()` because the function can do both regression and classification. To make this more interesting, we'll switch back to the `airquality` data. From the model summary below, `SVM-type:  eps-regression` tells us that the function is performing regression and not classification, then we see the hyperparameter values and the number of support vectors used to fit the model.

For the kernel, we have four choices: linear, polynomial, radial basis, and sigmoid. Selecting a linear kernel will force a straight line fit, and the other three kernels are different methods for adding curvature to the regression line^[Changing the kernel to specify the type of fit is known as the kernel trick.]. The theory behind SVR kernels is beyond the scope of this tutorial, but if you want to dig deeper:

* Here are some slides titled <a href="http://www.robots.ox.ac.uk/~az/lectures/ml/lect3.pdf">SVM dual, kernels and regression</a> from The University of Oxford.

* Here's <a href="http://web.mit.edu/6.034/wwwbob/svm-notes-long-08.pdf">An Idiot's Guide to Support Vector Machines</a>, a catchy title from MIT.

* Here's post titled <a href="https://towardsdatascience.com/understanding-support-vector-machine-part-2-kernel-trick-mercers-theorem-e1e6848c6c4d">Support Vector Machine: Kernel Trick; Mercer’s Theorem</a> at towardsdatascience.com. 

For our purposes, we just need to know that the three non-linear kernels have `gamma` as a hyperparameter that controls curvature. 

To force a straight regression line, specify `kernel='linear'`. Also, the `svm()` by default scales all variables in the data set to have a mean of zero and equal variance. Scaling the variables will improve the model's performance, but we'll turn that off in this example so we can directly compare the coefficients to those produced by `lm()`. 

```{r}
library(e1071)

aq.svm = svm(Ozone ~ Solar.R, data=aq, kernel='linear', scale=FALSE)
summary(aq.svm)
```

We can then extract the coefficients with `coef()`.

```{r}
(coeffs = coef(aq.svm))
```

Using `lm()`, we get the following coefficients.

```{r}
aq.lm = lm(Ozone ~ Solar.R, data=aq)
summary(aq.lm)
```

The coefficients produced by the two models might seem fairly different. The following plot shows the data with the two regression lines for comparison. Notice how the linear model is more influenced by the extreme high ozone values (possible outliers).

```{r}
ggplot() +
  geom_point(data = aq, aes(x=Solar.R, y=Ozone)) +
  geom_abline(slope=coeffs[2], intercept=coeffs[1], color='red') + 
  annotate("text", x=315, y=50, label="svm()", color='red') +
  geom_abline(slope=aq.lm$coefficients[2], intercept=aq.lm$coefficients[1], color='blue') +
  annotate("text", x=315, y=70, label="lm()", color='blue') +
  theme_bw()

```

Now we'll re-fit the model with a non-linear regression line and invoking scaling. To extract the predicted response, we use the `predict()` function just like with linear models. Plotting the predicted response gives is the following.

```{r}
aq.svm2 = svm(Ozone ~ Solar.R, data=aq)

aq = aq %>% mutate(svrY = predict(aq.svm2, data=aq))

ggplot(aq) +
  geom_point(aes(Solar.R, Ozone), color='black') +
  geom_line(aes(Solar.R, svrY), color='red') +
  ggtitle("SVR With Default Hyperparameters") +
  coord_fixed() +
  theme_bw()
 
```

To tune the hyperparameters with cross validation, we can use the `tune` function from the `e1017` package. If we give the `tune` function a range of values for the hyperparameters, it will perform a grid search of those values. In the following example, we're therefore fitting 100 different models. If we print the object returned from `tune`, we see that it performed 10-fold cross validation, the best hyperparameter values, and the mean squared error of the best performing model.

```{r}
set.seed(42)
aq.tune = tune.svm(Ozone ~ Solar.R, data = aq, gamma=seq(0.1, 1, 0.1), cost = seq(1, 100, 10))
print(aq.tune)
```

We can visualize the tune results as well by printing the `aq.tune` object. Here we see the range of cost and epsilon values with their associated mean squared error. The lower the error, the better, and those are indicated by the darkest blue regions.

```{r}
plot(aq.tune)
```

I prefer to choose a wide range of tuning parameter values initially, and then do a finer search in the area with the lowest error. It looks like we need a low gamma and a high cost.

```{r}
set.seed(42)
aq.tune = tune.svm(Ozone ~ Solar.R, data = aq, gamma=seq(0.02, 0.22, 0.05), cost = seq(80, 100, 2))
print(aq.tune)
```

The best model from the tuning call can be obtained with `aq.tune$best.model`, and we can then apply the `predict` function to get the best fit regression.

```{r}
aq$svrY = predict(aq.tune$best.model, data=aq)

ggplot(aq) +
  geom_point(aes(Solar.R, Ozone), color='black') +
  geom_line(aes(Solar.R, svrY), color='red') +
  ggtitle("SVR With Tuned Hyperparameters") +
  coord_fixed() +
  theme_bw()
```

### Support Vector Classification

Classification problems have either a binary or categorical response variable. To demonstrate how SVC works, we'll start with the `iris` data set, which contains four predictors and one categorical response variable. Plotting petal length versus petal width for the setosa and versicolor species shows that the two species are **linearly separable**, meaning we can draw a straight line on the plot that completely separates the two species. If we want to train an SVC to make predictions on new data, the question becomes: how do we draw the line that separates the data? There are infinitely many options, three of which are shown on the plot.

```{r echo=FALSE}
ggplot(iris %>% filter(Species %in% c('setosa', 'versicolor')), aes(x=Petal.Length, y=Petal.Width, color=Species)) +
  geom_point() +
  geom_abline(slope=1.2, intercept=-2.2, color='blue', linetype=2, size=1) +
  geom_abline(slope=0, intercept=0.8, color='red', linetype=3, size=1) +
  geom_abline(slope=-1, intercept=3.1, color='green', linetype=4, size=1) +
  theme_bw()
```

Support vector classification uses margins, but in a different way than SVR, to find a line that separates the data. If you think of the two parallel margin lines as a street, the idea is that we want to fit the widest possible street between the species because doing so results in the rest of the data points being as far off the street as possible. The two points below that fall on the margin determine the location of the support vectors.

```{r echo=FALSE}
ggplot(iris %>% filter(Species %in% c('setosa', 'versicolor')), aes(x=Petal.Length, y=Petal.Width, color=Species)) +
  geom_abline(slope=-1, intercept=3.2, color='black', linetype=1, size=1) +
  geom_abline(slope=-1, intercept=2.3, color='black', linetype=2, size=1) +
  geom_abline(slope=-1, intercept=4.1, color='black', linetype=2, size=1) +
  geom_point(size=2) +
  theme_bw()
```

What happens when two categories aren't linearly separable, as is the case when we look at versicolor and virginica below? 

```{r echo=FALSE}
iris_vv = droplevels(iris %>% filter(Species %in% c('versicolor', 'virginica')))

ggplot(iris_vv, aes(x=Petal.Length, y=Petal.Width, color=Species)) +
  geom_point() +
  theme_bw()
```

We still want to draw two parallel lines through the data sets, but the only way to do it is to have some observations in the middle of the street, or even on the wrong side of the line (called **margin violations**). We still want to fit as wide of a street as possible through the data points, but now we must also limit the number of margin violations. As with SVR, we can assign a **cost** for each margin violation. Since margin violations are generally bad, we might be tempted to apply a large cost; however, we must also consider how well the model will generalize. Below are the linear boundaries for two choices of cost. Support vectors are based on the points surrounded by black.

```{r echo=FALSE, fig.width=8, message=FALSE, warning=FALSE}
iris.svm = svm(Species~ Petal.Length + Petal.Width, kernel='linear', cost=1, scale=FALSE,
    data = iris_vv)

w = t(iris.svm$coefs) %*% iris.svm$SV # weight vector
slope = -w[1]/w[2]
intercept = iris.svm$rho/w[2]

df_sv = iris_vv[iris.svm$index, ] # get the support vectors

p1 = ggplot(iris_vv, aes(x=Petal.Length, y=Petal.Width, color=Species)) +
  geom_point(data=df_sv, aes(x=Petal.Length, y=Petal.Width), color='black', size = 5, alpha=0.5) +
  geom_abline(slope=slope, intercept=intercept, color='black', linetype=1, size=1) +
  geom_abline(slope=slope, intercept=intercept+1/w[2], color='black', linetype=2, size=1) +
  geom_abline(slope=slope, intercept=intercept-1/w[2], color='black', linetype=2, size=1) +
  geom_point() +
  theme_bw()

iris.svm2 = svm(Species~ Petal.Length + Petal.Width, kernel='linear', cost=100, scale=FALSE,
    data = iris_vv)

w2 = t(iris.svm2$coefs) %*% iris.svm2$SV # weight vector
slope2 = -w2[1]/w2[2]
intercept2 = iris.svm2$rho/w2[2]

df_sv2 = iris_vv[iris.svm2$index, ] # get the support vectors

p2 = ggplot(iris_vv, aes(x=Petal.Length, y=Petal.Width, color=Species)) +
  geom_point(data=df_sv2, aes(x=Petal.Length, y=Petal.Width), color='black', size = 5, alpha=0.5) +
  geom_abline(slope=slope2, intercept=intercept2, color='black', linetype=1, size=1) +
  geom_abline(slope=slope2, intercept=intercept2+1/w2[2], color='black', linetype=2, size=1) +
  geom_abline(slope=slope2, intercept=intercept2-1/w2[2], color='black', linetype=2, size=1) +
  geom_point() +
  theme_bw()

GGally::ggmatrix(list(p1, p2), nrow=1, ncol=2, xAxisLabels = c("Cost = 1", "Cost = 100"), legend=c(1,1))
```

Interestingly, the margins (and therefore the decision boundary) don't have to be straight lines. SVC also accommodates a curved boundary as in the example below. With a polynomial kernel, the curvature is controlled by the degree of the polynomial. In the plot, note that the support vectors are the `X` points.

```{r echo=FALSE}
iris.svm3 = svm(Species~ Petal.Length + Petal.Width, kernel='polynomial', degree=4, cost=1, scale=FALSE,
    data = iris_vv)
plot(iris.svm3, Petal.Width ~ Petal.Length, data=iris_vv)
```

#### Example In *R*

In this section, we'll walk through an example using the full `iris` data set. First, we'll split the data set into a training set that includes 80% of the data, and a test set with the remaining 20% using the `caTools` package.

```{r}
set.seed(0)
train = caTools::sample.split(iris, SplitRatio = 0.8)
iris_train = subset(iris, train == TRUE)
iris_test = subset(iris, train == FALSE)
```

Next, we'll tune two models using a linear kernel and a radial basis function (which allows for curvature). We'll tune both models over a range of gamma and cost values.

```{r}
iris.lin = tune.svm(Species~., data=iris_train, 
                    kernel="linear", 
                    gamma = seq(0.1, 1, 0.1), 
                    cost = seq(1, 100, 10))

iris.rbf = tune.svm(Species~., data=iris_train, 
                    kernel="radial", 
                    gamma = seq(0.1, 1, 0.1), 
                    cost = seq(1, 100, 10))

iris.lin$best.model
iris.rbf$best.model
```

Both models are using a low cost, but the radial basis function model has twice as many support vectors. To compare model performance, we'll make predictions using the test set and display each model's **confusion matrix** using the `cvms` package (note: we could also create a simple confusion matrix with `table(iris_test[, 5], predictions)`).

```{r message=FALSE, warning=FALSE}
# get the confusion matrix for the linear kernel
lin_conf_mat = cvms::confusion_matrix(
  targets = iris_test[, 5], 
  predictions = predict(iris.lin$best.model, type = 'response', newdata = iris_test[-5]))

# get the confusion matrix for the radial kernel
rbf_conf_mat = cvms::confusion_matrix(
  targets = iris_test[, 5],
  predictions = predict(iris.rbf$best.model, type = 'response', newdata = iris_test[-5]))

# plot the confusion matrix for the linear kernel (it's a ggplot2 object!)
cvms::plot_confusion_matrix(lin_conf_mat$`Confusion Matrix`[[1]]) + ggtitle("Linear Kernel")
```

The SVC model with the linear kernel did a great job! Of the 30 observations in the test set, only two were incorrectly classified. If this is the first time you've seen a confusion matrix, then what you see are the target (or actual) species by column and the species predictions from the SVC by row. In each cell, we see the percent and count of the total observations that fell into that cell. From this plot, we can identify true positives, false positives, etc. using the following guide.

$~$

<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;margin:0px auto;}
.tg td{border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;
  padding:10px 5px;word-break:normal;}
.tg th{border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;font-weight:normal;
  overflow:hidden;padding:10px 5px;word-break:normal;}
.tg .tg-18eh{border-color:#000000;font-weight:bold;text-align:center;vertical-align:middle}
.tg .tg-wp8o{border-color:#000000;text-align:center;vertical-align:top}
.tg .tg-xs2q{border-color:#000000;font-size:medium;text-align:center;vertical-align:middle}
.tg .tg-mqa1{border-color:#000000;font-weight:bold;text-align:center;vertical-align:top}
.tg .tg-1tol{border-color:#000000;font-weight:bold;text-align:left;vertical-align:middle}
</style>
<table class="tg" style="undefined;table-layout: fixed; width: 354px">
<colgroup>
<col style="width: 109px">
<col style="width: 54px">
<col style="width: 95px">
<col style="width: 95px">
</colgroup>
<tbody>
  <tr>
    <td class="tg-xs2q" colspan="2" rowspan="2"><span style="font-weight:bold">Confusion</span><br><span style="font-weight:bold">Matrix</span></td>
    <td class="tg-mqa1" colspan="2">Target</td>
  </tr>
  <tr>
    <td class="tg-mqa1">Yes</td>
    <td class="tg-mqa1">No</td>
  </tr>
  <tr>
    <td class="tg-1tol" rowspan="2">Prediction</td>
    <td class="tg-18eh">Yes</td>
    <td class="tg-wp8o">True<br>Positive</td>
    <td class="tg-wp8o">False<br>Positive</td>
  </tr>
  <tr>
    <td class="tg-18eh">No</td>
    <td class="tg-wp8o">False<br>Negative</td>
    <td class="tg-wp8o">True<br>Positive</td>
  </tr>
</tbody>
</table>

$~$

A perfect classifier will have zeros everywhere in the table except the diagonal. In our case, it's close to perfect. We just have two false negatives because two flowers that were actually virginica, were predicted to be versicolor. Now let's look at the radial kernel results.

```{r message=FALSE, warning=FALSE}
cvms::plot_confusion_matrix(rbf_conf_mat$`Confusion Matrix`[[1]]) + ggtitle("Radial Kernel")
```

### SVM Problem Set 

The problem set for this section is located <a href = '/_Chapter11_ProblemSets/SVM_PS_Questions.html'>here</a>.

For your convenience, the R markdown version is <a href = '/_Chapter11_ProblemSets/SVM_PS_Questions.Rmd'>here</a>.

The solutions are located <a href = '/_Chapter11_ProblemSets/SVM_PS_Solutions.html'>here</a>.

## Classification and Regression Trees

As with support vector machines, and as the name implies, classification and regression trees^[Sometimes referred to as partition trees.] (CART) can be used for either classification or regression tasks. Again, we'll start with regression and then move to classification. 

### Regression Trees

The algorithm is best explained as we walk through an example, and we'll continue to use the `airquality` data set. The basic machine learning algorithm used in tree-based methods follows these steps:

1. Consider the entire data set including all predictors and the response. We call this the **root node**, and it is represented by the top center node in the figure below. The information displayed in the node includes the mean response for that node (42.1 is the mean of `Ozone` for the whole data set), the number of observations in the node (`n=116`), and the percent of the overall observations in the node.

2. Iterate through each predictor, $k$, and split the data into two subsets (referred to as the left and right **child nodes**) using some threshold, $t_k$. For example, with the `airquality` data set, the predictor and threshold could be `Temp >= 83`. The choice of $k$ and $t_k$ for a given split is the pair that increases the "purity" of the child nodes (weighted by their size) the most. We'll explicitly define purity shortly. If you equate a data split with a decision, then at this point, we have a basic decision tree. 

```{r echo=FALSE}
library(rpart)
library(rpart.plot)

aq.tree = rpart(Ozone~., airquality, cp=0.4)
#printcp(aq.tree)
rpart.plot(aq.tree, digits = 3, type=4, extra=101)
```

3. Each child node in turn becomes the new parent node and the process is repeated. Below is the decision tree produced by the first two splits. Notice that the first split is on the `Temp` predictor, and the second split is on the `Wind` predictor. Although we don't have coefficients for these two predictors like we would in a linear model, we can still interpret the order of the splits as the predictor's relative significance. In this case, `Temp` is the most significant predictor of `Ozone` followed by `Wind`. After two splits, the decision tree has three **leaf nodes**, which are those in the bottom row. We can also define the **depth** of the tree as the number rows in the tree below the root node (in this case depth = 2). Note that the sum of the observations in the leaf nodes equals the total number of observations (69 + 10 + 37 = 116), and so the percentages shown in the leaf nodes sum to 100%.

```{r echo=FALSE}
aq.tree = rpart(Ozone~., airquality, cp=0.06)
rpart.plot(aq.tree, digits = 3, type=4, extra=101)
```

4. Continuing the process once more, we see that the third split is again on `Temp` but at a different $t_k$. 

```{r echo=FALSE}
aq.tree = rpart(Ozone~., airquality, cp=0.04)
rpart.plot(aq.tree, digits = 3, type=4, extra=101)
```

If we continued to repeat the process until each observation was in its own node, then we would have drastically over-fit the model. To control over-fitting, we stop the splitting process when some user-defined condition (or set of conditions) is met. Example stopping conditions include a minimum number of observations in a node or a maximum depth of the tree. We can also use cross validation with a 1 standard error rule to limit the complexity of the final model.  

We'll stop at this point and visually represent this model as a scatter plot. The above leaves from left to right are labeled as Leaf 1 - 4 on the scatter plot.

```{r echo=FALSE}
ggplot(airquality, aes(x=Wind, y = Temp)) +
  geom_point(aes(color=Ozone)) +
  geom_segment(aes(x=1, xend=21, y=82.5, yend=82.5, linetype="First Split"), color='blue', size=1) +
  geom_segment(aes(x=7.15, xend=7.15, y=55, yend=82.5, linetype="Second Split"), color='blue', size=1) +
  geom_segment(aes(x=1, xend=21, y=87.5, yend=87.5, linetype="Third Split" ), color='blue', size=1) +
  annotate("label", x=3.5, y=65, label="Leaf 2\nmean(Ozone)=55.6", color='blue') +
  annotate("label", x=16, y=65, label="Leaf 1\nmean(Ozone)=22.3", color='blue') +
  annotate("label", x=15, y=85, label="Leaf 3 mean(Ozone)=63.0", color='blue') +
  annotate("label", x=15, y=95, label="Leaf 4\nmean(Ozone)=90.1", color='blue') +
  scale_linetype_manual(name="Legend", values=c(1,2,3)) +
  ggtitle("Regression Tree Partitions") +
  theme_bw()
```

Plotting predicted `Ozone` on the z-axis produces the following response surface, which highlights the step-like characteristic of regression tree predictions.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(plotly)

tree_surface = expand_grid(Wind=0:21, Temp=57:97)
tree_surface = tree_surface %>% mutate(Solar.R = 0, Month = 0, Day = 0)
z = matrix(predict(aq.tree, newdata=tree_surface), byrow=F, ncol=22)

plot_ly() %>%
  add_trace(x=0:21, y=57:97, z = ~z, type='surface', opacity=0.9, showlegend=FALSE) %>%
  add_trace(data = aq, x=~Wind, y = ~Temp, z = ~Ozone, type='scatter3d', mode='markers', 
            marker=list(color='black', size=3), showlegend=FALSE) %>%
  layout(title = 'CART Response Surface', showlegend = FALSE)
```

Plotting just `Temp` versus `Ozone` in two dimensions further highlights a difference between this method and linear regression. From this plot we can infer that linear regression may outperform CART if there is a smooth trend in the relationship between the predictors and response because CART does not produce smooth estimates. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(airquality, aes(x=Temp, y = Ozone)) +
  geom_point() +
  geom_smooth(aes(color='Linear Regression'), method=lm, se=FALSE) +
  geom_segment(aes(x=55, xend=83, y=22.3, yend=22.3, color='CART'), size=1) +
  geom_segment(aes(x=83, xend=88, y=63, yend=63), color='blue', size=1) +
  geom_segment(aes(x=88, xend=97, y=90.1, yend=90.1), color='blue', size=1) +
  scale_colour_manual(name="Legend", values=c("blue", "red")) +
  ggtitle("CART vs. Linear Regression") +
  theme_bw()
```

#### Impurity Measure

Previously, it was stated that the predictor-threshold pair chosen for a split was the pair that most increased the purity (or, decreased the impurity) of the child nodes. A node with all identical response values will have an impurity of 0, so that as a node becomes more impure, it's impurity value increases. We will then define a node's impurity to be proportional to the **residual deviance**, which for a continuous response variable like `Ozone`, is the residual sum of squares (RSS).

$$RSS = \sum\limits_{i\:in\: Node}{(y_{i} - \bar{y})^2}$$

where $\bar{y}$ is the mean of the y's in the node.

We'll start with the first split. To determine which predictor-threshold pair decreases impurity the most, start with the first factor, send the lowest `Ozone` value to the left node and the remainder to the right node, and calculate RSS for each child node ($RSS_{left}$ and $RSS_{right}$). The decrease in impurity for this split is $RSS_{root} - (RSS_{left} + RSS_{right})$. Then send the lowest two `Ozone` values to the left node and the remainder to the right. Repeat this process for each predictor-threshold pair, and split the data based using the pair that decreased impurity the most. Any regression tree package will iterate through all of these combinations for you, but to demonstrate the process explicitly, We'll just consider the `Temp` predictor for the first split.

```{r}
# we'll do a lot of filtering, so convert dataframe to tibble for convenience
# we'll also drop the NA's for the calculations (but the regression tree
# methodology itself doesn't care if there are NA's or not)
aq  = as_tibble(airquality) %>% drop_na(Ozone)

# root node deviance
root_dev = sum((aq$Ozone - mean(aq$Ozone))^2) 

# keep track of the highest decrease
best_split = 0

# iterate through all the unique Temp values
for(s in sort(unique(aq$Temp))){
  left_node = aq %>% filter(Temp <= s) %>% .$Ozone
  left_dev = sum((left_node - mean(left_node))^2)
  right_node = aq %>% filter(Temp > s) %>% .$Ozone
  right_dev = sum((right_node - mean(right_node))^2)
  split_dev = root_dev - (left_dev + right_dev)
  if(split_dev > best_split){
    best_split = split_dev
    temp = s + 1}  # + 1 because we filtered Temp <= s and Temp is integer
}

print(paste("Best split at Temp <", temp), quote=FALSE)
```

#### Tree Deviance

Armed with our impurity measure, we can also calculate the tree deviance, which we'll use to calculate the regression tree equivalent of $R^2$. For the tree with the four leaf nodes, we calculate the deviance for each leaf. 

```{r}
# leaf 1
leaf_1 = aq %>% filter(Temp < 83 & Wind >= 7.15) %>% .$Ozone
leaf_1_dev = sum((leaf_1 - mean(leaf_1))^2)
# leaf 2
leaf_2 = aq %>% filter(Temp < 83 & Wind < 7.15) %>% .$Ozone
leaf_2_dev = sum((leaf_2 - mean(leaf_2))^2)
# leaf 3
leaf_3 = aq %>% filter(Temp >= 83 & Temp < 88) %>% drop_na(Ozone) %>% .$Ozone
leaf_3_dev = sum((leaf_3 - mean(leaf_3))^2)
# leaf 4
leaf_4 = aq %>% filter(Temp >= 88) %>% drop_na(Ozone) %>% .$Ozone
leaf_4_dev = sum((leaf_4 - mean(leaf_4))^2)
```

The tree deviance is the sum of the leaf node deviances, which we use to determine how much the entire tree decreases the root deviance.

```{r}
tree_dev = sum(leaf_1_dev, leaf_2_dev, leaf_3_dev, leaf_4_dev)

(root_dev - tree_dev) / root_dev
```

The tree decreases the root deviance by 61.2%, which also means that 61.2% of the variability in `Ozone` is explained by the tree.

#### Prediction

Making a prediction with a new value is easy as following the logic of the decision tree until you end up in a leaf node. The mean of the response values for that leaf node is the prediction for the new value.

#### Pros And Cons

Regression trees have a lot of good things going for them:

* They are easy to explain combined with an intuitive graphic output
* They can handle categorical and numeric predictor and response variables
* They easily handle missing data
* They are robust to outliers
* They make no assumptions about normality
* They can accommodate "wide" data (more predictors than observations)
* They automatically include interactions

Regression trees by themselves and as presented so far have two major drawbacks:

* They do not tend to perform as well as other methods (but there's a plan for this that makes them one of the best prediction methods around)
* They do not capture simple additive structure (there's a plan for this, too)

#### Regression Trees in *R*

The regression trees shown above were grown using the `rpart` and `rpart.plot` packages. I didn't show the code so that we could focus on the algorithm first. Growing a regression tree is as easy as a linear model. The object created by `rpart()` contains some useful information.

```{r}
library(rpart)
library(rpart.plot)

aq.tree = rpart(Ozone ~ ., data=airquality)

aq.tree
```

First, we see that the NAs were deleted, and then we see the tree structure in a text format that includes the node number, how the node was split, the number of observations in the node, the deviance, and the mean response. To plot the tree, use `rpart.plot()` or `prp()`.

```{r}
rpart.plot(aq.tree)
```

`rpart.plot()` provides several options for customizing the plot, among them are `digits`, `type`, and `extra`, which I invoked to produce the earlier plots. Refer to the help to see all of the options.

```{r}
rpart.plot(aq.tree, digits = 3, type=4, extra=101)
```

Another useful function is `printcp()`, which provides a deeper glimpse into what's going on in the algorithm. Here we see that just three predictors were used to grow the tree (`Solar.R`, `Temp`, and `Wind`). This means that the other predictors did not significantly contribute to increasing node purity, which is equivalent to a predictor in a linear model with a high p-value. We also see the root node error (weighted by the number of observations in the root node). 

In the table, `printcp()` provides optimal tuning based on a **complexity parameter** (`CP`), which  we can manipulate to manually "prune" the tree, if desired. The relative error column is the amount of reduction in root deviance for each split. For example, in our earlier example with three splits and four leaf nodes, we had a 61.2% reduction in root deviance, and below we see that at an `nsplit` of 3, we also get $1.000 - 0.388 = 61.2$%.^[It's always nice to see that I didn't mess up the manual calculations.] `xerror` and `xstd` are cross-validation error and standard deviation, respectfully, so we get cross validation built-in for free! 

```{r}
printcp(aq.tree)
```

With `plotcp()` we can see the 1 standard error rule implemented in the same manner we've seen before to identify the best fit model. At the top of the plot, the number of splits is displayed so that we can choose two splits when defining the best fit model.

```{r}
plotcp(aq.tree, upper = "splits")
```

Specify the best fit model using the `cp` parameter with a value slightly greater than shown in the table.

```{r}
best_aq.tree = rpart(Ozone ~ ., cp=0.055, data=airquality)

rpart.plot(best_aq.tree)
```

As with `lm()` objects, the `summary()` function provides a wealth of information. Note the results following variable importance. Earlier we opined that the first split on `Temp` indicated that is was the most significant predictor followed by `Wind`. The `rpart` documentation provides a detailed description of variable importance:

>An overall measure of variable importance is the sum of the goodness of split measures for each split for which it was the primary variable, plus goodness * (adjusted agreement) for all splits in which it was a surrogate.

Note that the results are scaled so that they sum to 100, which is useful for directly comparing each predictor's relative contribution.

```{r}
summary(aq.tree)
```

The best fit model contains two predictors and explains 55.8% of the variance in `Ozone` as shown below.

```{r}
printcp(best_aq.tree)
```

How does it compare to a linear model with the same two predictors? The linear model explains 56.1% of the variance in `Ozone`, which is only slightly more than the regression tree. Earlier I claimed there was a plan for improving the performance of regression trees. That plan is revealed in the next section on Random Forests.

```{r}
summary(lm(Ozone~Wind + Temp, data=airquality))
```

### Random Forest Regression

In 1994, Leo Breiman at UC, Berkeley published <a href="https://www.stat.berkeley.edu/~breiman/bagging.pdf">this paper</a> in which he presented a method he called **Bootstrap AGGregation** (or BAGGing) that improves the predictive power of regression trees by growing many trees (a forest) using bootstrapping techniques (thereby making it a random forest). The details are explained in the link to the paper above, but in short, we grow many trees, each on a bootstrapped sample of the training set (i.e., sample $n$ times *with replacement* from a data set of size $n$). Then, to make a prediction, we either let each tree "vote" and predict based on the most votes, or we use the average of the estimated responses. Cross-validation isn't necessary with this method because each bootstrapped tree has an internal error, referred to as the **out-of-bag (OOB) error**. With this method, about a third of the samples are left out of the bootstrapped sample, a prediction is made, and the OOB error calculated. The algorithm stops when the OOB error begins to increase.  

A drawback of the method is that larger trees tend to be correlated with each other, and so <a href="https://www.stat.berkeley.edu/~breiman/randomforest2001.pdf">in a 2001 paper</a>, Breiman developed a method to lower the correlation between trees. For each bootstrapped sample, his idea was to use a random selection of predictors to split each node. The number of randomly selected predictors, **mtry**, is a function of the total number of predictors in the data set. For regression, the `randomForest()` function from the `randomForest` package uses $1/k$ as the default `mtry` value, but this can be manually specified. The following code chunks demonstrate the use of some of the `randomForest` functions. First, we fit a random forest model and specify that we want to assess the importance of predictors, omit `NA`s, and randomly sample two predictors at each split (`mtry`). There are a host of other parameters that can be specified, but we'll keep them all at their default settings for this example.

```{r message=FALSE}
library(randomForest)

set.seed(42)

aq.rf<- randomForest(Ozone~., importance=TRUE, na.action=na.omit, mtry=2, data=airquality)
aq.rf
```

This random forest model consists of 500 trees and explains 72.% of the variance in `Ozone`, which is a nice improvement over the 55.8% we got with the single regression tree. Plotting the `aq.rf` object shows the error as a function of the size of the forest. We want to see the error stabilize as the number of trees increases, which it does in the plot below.

```{r}
plot(aq.rf)
```

#### Interpretation

When the relationships between predictors and response are non-linear and complex, random forest models generally perform better than standard linear models. However, the increase in predictive power comes with a corresponding decrease in interpretability. For this reason, random forests and some other machine learning-based models such as neural networks are sometimes referred to as "black box" models. If you are applying machine learning techniques to build a model that performs optical character recognition, you might not be terribly concerned about the interpretability of your model. However, if your model will be used to inform a decision maker, interpretability is much more important - especially if you are asked to explain the model to the decision maker. In fact, some machine learning practitioners argue against using black box models for all high stakes decision making. For example, read <a href="https://arxiv.org/pdf/1811.10154.pdf">this paper</a> by Cynthia Rudin, a computer scientist at Duke University. Recently, advancements have been made in improving the interpretability of some types of machine learning models (for example, download and read <a href="https://www.h2o.ai/resources/ebook/introduction-to-machine-learning-interpretability/">this paper from h2o.ai</a> or <a href="https://christophm.github.io/interpretable-ml-book/">this e-book</a> by Christoph Molnar, a Ph.D. candidate at the University of Munich), and we will explore these techniques below. 

Linear models have coefficients (the $\beta$s) that explain the nature of the relationship between predictors and the response. Classification and regression trees have an analogous concept of variable importance, which can be extended to random forest models. The documentation for `importance()` from the `randomForest` package provides the following definitions of two variable importance measures:

>The first measure is computed from permuting OOB data: For each tree, the prediction error on the out-of-bag portion of the data is recorded (error rate for classification, MSE for regression). Then the same is done after permuting each predictor variable. The difference between the two are then averaged over all trees, and normalized by the standard deviation of the differences. If the standard deviation of the differences is equal to 0 for a variable, the division is not done (but the average is almost always equal to 0 in that case).

>The second measure is the total decrease in node impurities from splitting on the variable, averaged over all trees. For classification, the node impurity is measured by the Gini index. For regression, it is measured by residual sum of squares.

These two measures can be accessed with:

```{r}
importance(aq.rf)
```

Alternatively, we can plot variable importance with `varImpPlot()`. 

```{r}
varImpPlot(aq.rf)
```

Variable importance can be related to a linear model coefficient in that a large variable importance value is akin to a large coefficient value. However, it doesn't indicate whether the coefficient is positive or negative. For example, from the above plot, we see that `Temp` is an important predictor of `Ozone`, but we don't know if increasing temperatures result in increasing or decreasing ozone measurements (or if it's a non-linear relationship). **Partial dependence** plots (PDP) were developed to solve this problem, and they can be interpreted in the same way as a loess or spline smoother.

For the `airquality` data, one would expect that increasing temperatures would increase ozone concentrations, and that increasing wind speed would decrease ozone concentrations. The `partialPlot()` function provided with the `randomForest` package produces PDPs, but they are basic and difficult to customize. Instead, we'll use the `pdp ` package, which works nicely with `ggplot2` and includes a loess smoother (another option is the `iml` package - for interpretable machine learning - which we'll also explore).

```{r message=FALSE, warning=FALSE}
#library(pdp)

p3 = aq.rf %>%
  pdp::partial(pred.var = "Temp") %>%                        # from the pdp package
  autoplot(smooth = TRUE, ylab = expression(f(Temp))) +
  theme_bw() +
  ggtitle("Partial Dependence of Temp")

p4 = aq.rf %>%
  pdp::partial(pred.var = "Wind") %>%
  autoplot(smooth = TRUE, ylab = expression(f(Temp))) +
  theme_bw() +
  ggtitle("Partial Dependence of Wind")

gridExtra::grid.arrange(p3, p4, ncol=2)
```

Earlier, we produced a response surface plot based on a regression tree. Now we can produce a response surface based on the random forest model, which looks similar but more detailed. Specifying `chull = TRUE` (chull stands for convex hull) limits the plot to the range of values in the training data set, which prevents predictions being shown for regions in which there is no data. A 2D heat map and a 3D mesh are shown below. 

```{r}
# Compute partial dependence data for Wind and Temp
pd = pdp::partial(aq.rf, pred.var = c("Wind", "Temp"), chull = TRUE)

# Default PDP
pdp1 = pdp::plotPartial(pd)

# 3-D surface
pdp2 = pdp::plotPartial(pd, levelplot = FALSE, zlab = "Ozone",
                    screen = list(z = -20, x = -60))

gridExtra::grid.arrange(pdp1, pdp2, ncol=2)
```

The `iml` package was developed by Christoph Molnar, the Ph.D. candidate referred to earlier, and contains a number of useful functions to aid in model interpretation. In machine learning vernacular, predictors are commonly called features, so instead of variable importance, we'll get feature importance. With this package, we can calculate feature importance and produce PDPs as well, and a grid of partial dependence plots are shown below. Note the addition of a rug plot at the bottom of each subplot, which helps identify regions where observations are sparse and where the model might not perform as well.

```{r}
#library(iml) # for interpretable machine learning
#library(patchwork) # for arranging plots - similar to gridExtra

# iml doesn't like NAs, so we'll drop them from the data and re-fit the model
aq = airquality %>% drop_na()
aq.rf2 = randomForest(Ozone~., importance=TRUE, na.action=na.omit, mtry=2, data=aq)

# provide the random forest model, the features, and the response
predictor = iml::Predictor$new(aq.rf2, data = aq[, 2:6], y = aq$Ozone)

PDP = iml::FeatureEffects$new(predictor, method='pdp')
PDP$plot() & theme_bw()
```

PDPs show the average feature effect, but if we're interested in the effect for one or more individual observations, then an Individual Conditional Expectation (ICE) plot is useful. In the following plot, each black line represents one of the 111 observations in the data set, and the global partial dependence is shown in yellow. Since the individual lines are generally parallel, we can see that each individual observation follows the same general trend: increasing temperatures have little effect on ozone until around 76 degrees, at which point all observations increase. In the mid 80s, there are a few observations that have a decreasing trend while the majority continue to increase, which indicates temperature may be interacting with one or more other features. Generally speaking, however, since the individual lines are largely parallel, we can conclude that the partial dependence measure is a good representation of the whole data set. 

```{r}
ice = iml::FeatureEffect$new(predictor, feature = "Temp", method='pdp+ice') #center.at = min(aq$Temp))
ice$plot() + theme_bw()
```

One of the nice attributes of tree-based models is their ability to capture interactions. The interaction effects can be explicitly measured and plotted as shown below. The x-axis scale is the percent of variance explained by interaction for each feature, so `Wind`, `Temp`, and `Solar.R` all have more than 10% of their variance explained by an interaction.

```{r}
interact = iml::Interaction$new(predictor)
plot(interact) + theme_bw()
```

To identify what the feature is interacting with, just specify the feature name. For example, `Temp` interactions are shown below.

```{r}
interact = iml::Interaction$new(predictor, feature='Temp')
plot(interact) + theme_bw()
```

#### Predictions

Predictions for new data are made the usual way with `predict()`, which is demonstrated below using the first two rows of the `airquality` data set.

```{r}
predict(aq.rf, airquality[1:2, c(2:6)])
```


### Random Forest Classification

For a classification example, we'll skip over simple classification trees and jump straight to random forests. There is very little difference in syntax with the `randomForest()` function when performing classification instead of regression. For this demonstration, we'll use the `iris` data set so we can compare results with the SVC results. We'll use the same training and test sets as earlier.

```{r}
set.seed(0)
iris.rf <- randomForest(Species ~ ., data=iris_train, importance=TRUE)
print(iris.rf)
```

The model seems to have a little trouble distinguishing virginica from versicolor. The linear SVC misclassified two observations in the test set, and the radial SVC misclassified one. Before we see how the random forest does, let's make sure we grew enough trees. We can make a visual check by plotting the random forest object. 

```{r}
plot(iris.rf)
```

No issue there! Looks like 500 trees was plenty. Taking a look at variable importance shows that petal width and length are far more important than sepal width and length.

```{r}
varImpPlot(iris.rf)
```

Since the response variable is categorical with three levels, a little work is required to get partial dependence plots for each predictor-response combination. Below are the partial dependence plots for `Petal.Width` for each species. The relationship between petal width and species varies significantly based on the species, which is what makes petal width have a high variable importance.

```{r}
as_tibble(iris.rf %>%
  pdp::partial(pred.var = "Petal.Width", which.class=1) %>% # which.class refers to the factor level
  mutate(Species = levels(iris$Species)[1])) %>%
  bind_rows(as_tibble(iris.rf %>%
  pdp::partial(pred.var = "Petal.Width", which.class=2) %>%
  mutate(Species = levels(iris$Species)[2]))) %>%
  bind_rows(as_tibble(iris.rf %>%
  pdp::partial(pred.var = "Petal.Width", which.class=3) %>%
  mutate(Species = levels(iris$Species)[3]))) %>%
  ggplot() +
  geom_line(aes(x=Petal.Width, y=yhat, col=Species), size=1.5) +
  ggtitle("Partial Dependence of Petal.Width") +
  theme_bw()
```

Enough visualizing. Time to get the confusion matrix for the random forest model using the test set.

```{r message=FALSE, warning=FALSE}
# get the confusion matrix
rf_conf_mat = cvms::confusion_matrix(
  targets = iris_test[, 5],
  predictions = predict(iris.rf, newdata = iris_test[-5]))

# plot the confusion matrix
cvms::plot_confusion_matrix(rf_conf_mat$`Confusion Matrix`[[1]]) + ggtitle("Random Forest")
```

Two observations were misclassified just like with the linear SVC. Let's see if they're the same two observations.

```{r}
# the indices of the misclassified flowers from SVC
which(iris_test[, 5] != predict(iris.lin$best.model, newdata = iris_test[-5]))

# the indices of the misclassified flowers from random forest
which(iris_test[, 5] != predict(iris.rf, newdata = iris_test[-5]))
```

### CART Problem Set 

The problem set for this section is located <a href = '/_Chapter11_ProblemSets/RF_PS_Questions.html'>here</a>.

For your convenience, the R markdown version is <a href = '/_Chapter11_ProblemSets/RF_PS_Questions.Rmd'>here</a>.

The solutions are located <a href = '/_Chapter11_ProblemSets/RF_PS_Solutions.html'>here</a>.