---
title: "DOE Packages in R"
output: html_document
---

```{r setup-10, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# DOE Packages in *R* {-}

## Admin

```{r include=FALSE}
mySession = sessionInfo()
```

This chapter was published using the following software:

* `r mySession$R.version$version.string`.
* On `r mySession$platform` running `r mySession$running`.
* This appendix looks at the following packages and their versions:
    * `FrF2` version `r packageVersion('FrF2')`.

## FrF2

```{r}
tempFunc <- function(n){return(data.frame(nfac = FrF2::catlg[[n]]$nfac, res = FrF2::catlg[[n]]$res))}
catlg.summary <- bind_rows(lapply(X = 1:length(FrF2::catlg), tempFunc))
min.fac <- min(catlg.summary$nfac)
max.fac <- max(catlg.summary$nfac)
```

`FrF2`, "Fractional Factorial Designs with 2-Level Factors" is an *R* package by <a href = "http://prof.beuth-hochschule.de/groemping/software/versuchsplanung/"> Ulrike Groemping </a>.  It can produce designs for experiments with `r min.fac` to `r max.fac` factors at varying resolutions.  It (appears) to do this by using a catalog of designs and fitting a design based on the variety of input arguments.  Below is a summary of the available design by number of factors, the associated minimum and maximum resolutions, and the number of designs available:

```{r}
temp <- catlg.summary %>% 
  group_by(nfac) %>% 
  summarise(Min.Available.Resolution = min(res), 
            Max.Available.Resolution = max(res), 
            Number.Designs.Available = n()) 

temp %>% ungroup() %>% group_by(Max.Available.Resolution) %>% summarize(nfac = max(nfac)) %>% arrange(nfac) %>% filter(nfac > 3) %>% select(nfac, Max.Available.Resolution)
```

Of note, for factors up to 65, we can get resolution 5.  For factors 66 - 160, we can get resolution 4

Each design can be accessed through the `FrF2` list `catlg`:

```{r}
FrF2::catlg
```

We can access an individual design either by index or by its name:

```{r}
FrF2::catlg$`3-1.1`

FrF2::catlg[[1]]
```

We can further access individual elements of this list in the usual way, e.g. `catlg[[1]]$gen` yields the generating column.

### Generating a Design

The `FrF2` command `FrF2` generates commands.  It has several possible inputs:

* nruns
* nfactors
* factor.names
* ncenter
* center.distribute
* generators
* design
* resolution
* 


```{r}
myDesign <- FrF2(nruns = 4, nfactors = 3, factor.names = c('Factor1', 'Factor2', 'Factor3'))
myDesign

generators(myDesign)
```

### Introduction

### Capabilities

### Producing a Design

### 
