---
title: "Cross Validation Problem Set"
author: 'Your Name Here'
output: 
  html_document:
    number_sections: T
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This problem set corresponds to the chapter on model selection in the course https://doe.dscoe.org. For this problem set, we will use the following packages:

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(bestglm)
library(glmnet)
```

As with the previous problem set , we will use the `SAheart` dataset from the `bestglm` package. As a reminder, from the help documentation, we find that the data consists of 462 observations of 10 variables. The response is `chd`, which is a binary variable indicating the presence (1) or absence (0) of coronary heart disease. 

# Train and Test Sets

Create training and test sets using an 80/20 split.

```{r}
set.seed(0)
test_idx = sample(462, 92, replace=FALSE)

train_set = SAheart[-test_idx, ]
test_set = SAheart[test_idx, ]
```

# AIC Model

Find the best model using AIC (remember that the response is binary). This should be the same model fit in the previous problem set.

```{r}
sa.glm = bestglm(train_set, IC="AIC", family=binomial)
sa.glm$BestModel
```

# `bestglm` Model

Find the best model using The "HTF" method with your choice for the number folds and replications. For reproducibility, choose a seed number before performing cross validation.

```{r}
set.seed(0)
sa.cv = bestglm(train_set, IC="CV", CVArgs=list(Method="HTF",K=10,REP=2), family=binomial)
sa.cv
```

## Predictors

Plot the number of predictors on the x-axis and the CV prediction error and the standard error on the y-axis. How many predictors are included in the model with the lowest CV error versus the best model?

```{r}
sa.df = tibble(
  cv = sa.cv$Subsets$CV,    # cross validation errors
  se = sa.cv$Subsets$sdCV,  # standard errors
  cv.low = cv - se,
  cv.high= cv + se,
  p = 0:(length(cv)-1)      # number of predictors
)

ggplot(sa.df) +
  geom_segment(aes(x=0:9, xend=0:9, y=cv.low, yend=cv.high), color='blue') +
  geom_line(aes(x=p, y=cv), color='red') +
  geom_point(aes(x=p, y=cv), color='red', size=3) +
  geom_hline(yintercept = sa.df %>% filter(p==6) %>% .$cv.high, color='blue', linetype=2) +
  scale_x_continuous(breaks = 0:9) +
  xlab("Number of Predictors") +
  ylab("CV Prediction Error") +
  theme_bw()
```

# Lasso Regression

Staying with the `SAheart` data, perform lasso regression, and plot log(lambda) vs. coefficients. Remember that the predictors and response variable need to be in separate data structures, and that the predictors should be in a `data.matrix`.

```{r}
x = data.matrix(train_set[-10])
y = train_set$chd

sa.lasso = glmnet(x, y, family="binomial")
plot(sa.lasso, xvar="lambda", label=TRUE)
```

## First Non-Zero Predictor

Which predictor had the first non-zero coefficient? Is it the same predictor that had the lowest p-value from the other methods?

```{r}
coef(sa.lasso, s=sa.lasso$lambda[2])
# age is first and also had the lowest p-value above
```

## Best Fit Model

Perform lasso regression with cross validation and print the coefficients for the best fit model. How does this model compare to the other methods?

```{r}
set.seed(0)
sa.lasso.cv = cv.glmnet(x, y, family="binomial")
coef(sa.lasso.cv, s = "lambda.1se")
# 4 factors - fewer than AIC, and more than CV
```

# Prediction Error

Compare the mean absolute error for the AIC, CV, and lasso models.

```{r}
# AIC
aic_preds = predict(sa.glm$BestModel, newdata = data.frame(test_set[, -10]))

# CV
cv.glm = glm(chd ~ tobacco + famhist + age, data=train_set)
cv_preds = predict(cv.glm, newdata = data.frame(test_set[, -10]))

# Lasso
lasso.glm = glm(chd ~ tobacco + ldl + famhist + age, data=train_set)
lasso_preds = predict(lasso.glm, newdata = data.frame(test_set[, -10]))

# calculate and compare mean absolute error
print(paste("AIC mean absolute error:", round(mean(abs(aic_preds - test_set$chd)), 3)))
print(paste("CV mean absolute error:", round(mean(abs(cv_preds - test_set$chd)), 3)))
print(paste("Lasso mean absolute error:", round(mean(abs(lasso_preds - test_set$chd)), 3)))
```

# Recommendation

Show the model summary of the model you would recommend using to make new predictions and explain why you chose it?

```{r}
summary(glm(chd ~ tobacco + ldl + famhist + age, data=SAheart, family=binomial))

# I'd recommend the lasso model because it incorporates cross validation, making 
# it more robust for predictive purposes as evidenced by it's relatively
# low mean absolute error.
```