---
title: "Cross Validation Problem Set"
author: 'Your Name Here'
output: 
  html_document:
    number_sections: T
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This problem set corresponds to the chapter on model selection in the course https://doe.dscoe.org. For this problem set, we will use the following packages:

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(bestglm)
library(glmnet)
```

As with the previous problem set , we will use the `SAheart` dataset from the `bestglm` package. As a reminder, from the help documentation, we find that the data consists of 462 observations of 10 variables. The response is `chd`, which is a binary variable indicating the presence (1) or absence (0) of coronary heart disease. 

# Train and Test Sets

Create training and test sets using an 80/20 split.

```{r}
# Answer here.
```

# AIC Model

Find the best model using AIC (remember that the response is binary). This should be the same model fit in the previous problem set.

```{r}
# Answer here.
```

# `bestglm` Model

Find the best model using The "HTF" method with your choice for the number folds and replications. For reproducibility, choose a seed number before performing cross validation.

```{r}
# Answer here.
```

## Predictors

Plot the number of predictors on the x-axis and the CV prediction error and the standard error on the y-axis. How many predictors are included in the model with the lowest CV error versus the best model?

```{r}
# Answer here.
```

# Lasso Regression

Staying with the `SAheart` data, perform lasso regression, and plot log(lambda) vs. coefficients. Remember that the predictors and response variable need to be in separate data structures, and that the predictors should be in a `data.matrix`.

```{r}
# Answer here.
```

## First Non-Zero Predictor

Which predictor had the first non-zero coefficient? Is it the same predictor that had the lowest p-value from the other methods?

```{r}
# Answer here.
```

## Best Fit Model

Perform lasso regression with cross validation and print the coefficients for the best fit model. How does this model compare to the other methods?

```{r}
# Answer here.
```

# Prediction Error

Compare the mean absolute error for the AIC, CV, and lasso models.

```{r}
# Answer here.
```

# Recommendation

Show the model summary of the model you would recommend using to make new predictions and explain why you chose it?

```{r}
# Answer here.
```