---
title: "Chapter 2: Descriptive Statistics - Problem Set"
author: "Your Name Here"
date: "Your Date Here"
output: 
  html_document:
    number_sections: true
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This tutorial is intended to allow people to practice producing and interpreting descriptive statistics and figures.  We use the Ames, Iowa data set as described in: http://jse.amstat.org/v19n3/decock.pdf.  The actual data set is contained in this folder for your convenience.

This tutorial assumes some base familiarity with ggplot and the tidyverse.  If you do not have that, please see previous lessons or google.
For building nice plots, the *R Graphics Cookbook* by Winston Chang is a useful reference: https://r-graphics.org/
More generally, if I don't know how to plot something, I google it and find the answer on someone's stackexchange or something like that.

For this tutorial, we will use the following packages:


```{r}
library(dplyr)
library(ggplot2)
```

# Step 1: Read the Data

We want to take the data from its source (in this case, a .csv) and convert it into an *R* object.


```{r}
# One useful way to do this is with the rio package
ames <- rio::import('./ames.csv')

# One can also use read.csv or read.csv2 to read ames data
ames.read.csv <- read.csv2('./ames.csv', sep = ',', stringsAsFactors = F)

# Both results are valid, note that there are differing defaults in how strings are read with differing commands.  For convenience, I'm going to remove ames.read.csv
rm(ames.read.csv)
```

## Get an Understanding of the Data:

1. Look at the top several rows of the data (hint head()).


```{r}
# head() is a nice way to look at the top several rows of a data set.  It gives you a quick idea as to what you're looking at
head(ames)
```

2. Find out how many fields (columns) and records (rows) exist in the data. (Note, there are many ways to do this).


```{r}
# Number of records (rows)
nrow(ames)

# Number of fields (columns)
ncol(ames)

# Alternative; you just have to know that dim produces rows then columns when the object passed is a data.frame
dim(ames)
```

3. Identify what the fields are and learn something about them.


```{r}
# One way to do this is to look at the columnnames
colnames(ames)

# Another way to do this is to use the str() command.  It shows you both the columns and their datatypes
str(ames)

# Finally, summary is useful, particularly when you have numeric data.
summary(ames)
```

4. Determine the number of rows that are missing at least one data record (i.e., the number of rows with an NA).


```{r}
# First, consider what it means for a data point to be an NA
head(is.na(ames)) # we only want to look at the first few rows of this output, so we wrap the query in head().  We can also change the global  option: options(max.print = N) where N is the number of rows you want to print

# Second you can sum this by rows (recall if you sum boolean values, T is given the value 1 and false is given the value 0, NA is given the value NA)
# the rowSums function is a nice one
head(rowSums(is.na(ames)))

# This does not tell you the total number of rows that have NAs, so we can summarize it:
# one way is to do a boolean test against the above set of numbers and then sum that test
sum(rowSums(is.na(ames)) > 0)

# A nice alternative is the summary function which tells us something about how the number of NAs are distributed across the rows
summary(rowSums(is.na(ames)))

# We can do a similar set of analysis with colSums()
colSums(is.na(ames))
```

At this point, you have to make some decision about your data (e.g., drop incomplete records, provide a standard fill to incomplete records, or something else).  We're not going to do that however.

## Select the appropriate data

Select the columns: "SalePrice" and "House.Style" and remove the rest.  We're only going to look at these for the rest of this section.


```{r}
# Option 1: Indexing
ames.base <- ames[,c('SalePrice', 'House.Style')]

# Option 2: dplyr
ames.dplyr <- dplyr::select(ames, SalePrice, House.Style)

# Option 3: dplyr with piping
library(dplyr)
ames.dplyr2 <- ames %>% select(SalePrice, House.Style)

# Since we only need one of these, I'll delete them all and just save over the ames data with only the two selected columns
rm(ames.base, ames.dplyr, ames.dplyr2)
ames <- dplyr::select(ames, SalePrice, House.Style)
```

Ensure that SalePrice is numeric and House.Style is a factor


```{r}
# We see SalePrice is an integer which is fine, but House.Style is a character, which we want to convert to a factor
str(ames)

# Convert ames$House.Style to a factor
ames$House.Style <- as.factor(ames$House.Style)

# Check out the difference
str(ames)
```

# Step 2. Continuous Data Descriptions

Find the maximum, minimum, median, mean, and standard deviation of the sale prices.


```{r}
# Maximum:
max(ames$SalePrice)

# Minimum
min(ames$SalePrice)

# Median
median(ames$SalePrice)

# Mean
mean(ames$SalePrice)

# Alternatively, one can get all of this very quickly using the summary function
summary(ames$SalePrice)

# Get the standard deviation of the data
sd(ames$SalePrice)
```

Build & plot a probability distribution (pdf) of the ames data.


```{r}
# The density function provides an approximation to the probability distribution function:
ames.density <- density(ames$SalePrice)

# The density function outputs an x value (in this case, the Sale Price)
head(ames.density$x)

# The density functino outputs a y value (in this case, the probability)
head(ames.density$y)

# You can plot this:
ggplot() + 
    geom_line(aes(x = ames.density$x, y = ames.density$y)) + 
    xlab('Housing Price ($)') + 
    ylab('Probability') + 
    ggtitle('Probability Distribution Function for Ames Sale Prices') 
```

Build an empirical cumulative density function, plot it, and find the probability that a house costs between $100,000 and $150,000


```{r}
ames.cdf <- ecdf(ames$SalePrice)

# Create a vector of costs
cost <- seq(0, 750000, by = 10000)
# Calculate their cumulative probability (percentile)
cum.prob <- ames.cdf(cost)
# Plot it
ggplot() + 
    geom_line(aes(x = cost, y = cum.prob)) + 
    ggtitle('CDF for Ames, Iowa House Sales') + 
    xlab('Cost ($)') + ylab('Percentile')

# You can assess the probability that a house will cost in a range by calculating the difference of the percentiles of the two numbers:
ames.cdf(150000) - ames.cdf(100000)
# It's about 35%
```

Find what the sale prices are for the 0, 10, 20, ... percentiles (i.e. the quantiles)


```{r}
# The quantile function is very useful.  You can pass it an arbitrary vector
quantile(ames$SalePrice, c(0, .1, .2, .3, .4, .5, .6, .7, .8, .9, 1))


# Note that ecdf and quantile are effectively inverses of each other (up to rounding errors)
quantile(ames$SalePrice, ecdf(ames$SalePrice)(100000))
```

Visualize the data using a boxplot, density plot, histogram


```{r}
# Box plot, note we are not varying by any other data type
ggplot(data = ames) + geom_boxplot(aes(x = '', y = SalePrice))

# Density plot (note this is the same as what we saw when we plotted the results of Density)
ggplot(data = ames) + geom_density(aes(x = SalePrice))

# Histogram
ggplot(data = ames) + geom_histogram(aes(x = SalePrice))
```

# Step 3. Categorical Data Description
For this section, note that House.Style is a categorical (in R, it is called a factor) variable.

What are the number of categories (levels) of the factor.  How many records are in each level


```{r}
# You can find the names of the levels for the variable with levels()
levels(ames$House.Style)

# Note that the summary function treats a factor different than a numeric data type.  It returns the levels and the number 
# of observations for each level
summary(ames$House.Style)
```

Plot the number of instances of each level in as a bar chart


```{r}
# Plot as a bar chart 
ggplot(data = ames) + geom_bar(aes(x = House.Style))
```

Plot the levels as a pie chart (this can be a bit tricky, I always end up googling how to do it)


```{r}
# It's helpful to turn this into a single data frame where the number of records for each level is recorded
ames.levels <- ames %>% group_by(House.Style) %>% summarize(N = n())
ggplot(data = ames.levels) + # we are using the ames.levels data we just created
    geom_bar(aes(x='', y = N, fill = House.Style), stat = 'identity') + # we're building a bar chart that: 
                                                                        # 1) has no x axis (everything on top of everything else), 
                                                                        # 2) uses the N column for its y axis, and 
                                                                        # 3) colors the fill of values based on the house style
    coord_polar('y', start = 0) # This changes the y coordinates from your normal linear to a polar type


ames.levels
```

# Step 4. Describe a mix of continuous and categorical data

Produce a boxplot chart where the x axis is the type of house


```{r}
ggplot(data = ames) + 
    geom_boxplot(aes(x = House.Style, y = SalePrice))
```

Produce multiple histograms where each chart is for a distinct house style


```{r}
ggplot(data = ames) + 
    geom_histogram(aes(x = SalePrice)) + 
    facet_wrap(~House.Style)
```

Produce a density plot where each house type is colored differently and overlayed on top of the other


```{r}
ggplot(data = ames) + 
    geom_density(aes(x = SalePrice, color = House.Style), fill = NA)
```

# Conclusion

At the end of this lesson you should be able to describe continuous and categorical data numerically and graphically.  There are many, many ways to describe data, this tutorial only covered a small sample.  I recommend you practice plotting in different ways using ggplot (or your other favorite plotting package, e.g., plotly, bokeh, or even base R).  If you
