---
title: "Chapter 2: Descriptive Statistics - Problem Set"
author: "Your Name Here"
date: "Your Date Here"
output: 
  html_document:
    number_sections: true
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```


# Introduction
This tutorial is intended to allow people to practice producing and interpreting descriptive statistics and figures.  We use the Ames, Iowa data set as described in: http://jse.amstat.org/v19n3/decock.pdf.  The actual data set is contained in this folder for your convenience.

This tutorial assumes some base familiarity with ggplot and the tidyverse.  If you do not have that, please see previous lessons or google.
For building nice plots, the *R Graphics Cookbook* by Winston Chang is a useful reference: https://r-graphics.org/
More generally, if I don't know how to plot something, I google it and find the answer on someone's stackexchange or something like that.

For this tutorial, we will use the following packages:


```{r}
library(dplyr)
library(ggplot2)
```

# Step 1: Read the Data

We want to take the data from its source (in this case, a .csv) and convert it into an *R* object.


```{r}
#Answer here

```

## Get an Understanding of the Data:

1. Look at the top several rows of the data (hint head()).


```{r}
#Answer here


```

2. Find out how many fields (columns) and records (rows) exist in the data. (Note, there are many ways to do this).


```{r}
#Answer here


```

3. Identify what the fields are and learn something about them.


```{r}
#Answer here


```

4. Determine the number of rows that are missing at least one data record (i.e., the number of rows with an NA).


```{r}
#Answer here


```

At this point, you have to make some decision about your data (e.g., drop incomplete records, provide a standard fill to incomplete records, or something else).  We're not going to do that however.

### Select the appropriate data

Select the columns: "SalePrice" and "House.Style" and remove the rest.  We're only going to look at these for the rest of this section.


```{r}
#Answer here


```

Ensure that SalePrice is numeric and House.Style is a factor


```{r}
#Answer here


```

# Step 2. Continuous Data Descriptions

Find the maximum, minimum, median, mean, and standard deviation of the sale prices.


```{r}
#Answer here


```

Build & plot a probability distribution (pdf) of the ames data.


```{r}
#Answer here


```

Build an empirical cumulative density function, plot it, and find the probability that a house costs between $100,000 and $150,000


```{r}
#Answer here


```

Find what the sale prices are for the 0, 10, 20, ... percentiles (i.e. the quantiles)


```{r}
#Answer here


```

Visualize the data using a boxplot, density plot, histogram


```{r}
#Answer here


```

# Step 3. Categorical Data Description

For this section, note that House.Style is a categorical (in R, it is called a factor) variable.

What are the number of categories (levels) of the factor?  How many records are in each level?


```{r}
#Answer here


```

Plot the number of instances of each level in as a bar chart.


```{r}
#Answer here


```

Plot the levels as a pie chart (this can be a bit tricky, I always end up googling how to do it).


```{r}
#Answer here


```

# Step 4. Describe a mix of continuous and categorical data

Produce a boxplot chart where the x axis is the type of house.


```{r}
#Answer here


```

Produce multiple histograms where each chart is for a distinct house style.


```{r}
#Answer here


```

Produce a density plot where each house type is colored differently and overlayed on top of the other.


```{r}
#Answer here


```

# Conclusion

At the end of this lesson you should be able to describe continuous and categorical data numerically and graphically.  There are many, many ways to describe data, this tutorial only covered a small sample.  I recommend you practice plotting in different ways using ggplot (or your other favorite plotting package, e.g., plotly, or even base R).  
